---
title: "Manuscript"
format: html
---

```{r}
#| label: dependancies
library(tidyverse)
library(sva)
library(ggpubr)
library(ggprism)
library(RColorBrewer)
library(broom)
library(rstatix)
source("01_preprocessing_helper_functions.R")
library(survival)
library(survminer)
library(patchwork)
library(phyloseq)
library(microbiome)
library(ggbeeswarm)
library(rstatix)
library(lefser)
library(ggupset)
library(MicrobiomeStat)
library(maaslin3)
library(glmnet)
library(pROC)
library(pheatmap)
library(vegan)
library(precrec)
```



# Preprocess Data

## Setup Output Directories

```{r}
outdir <- "../2_analysis/01_preprocessedData"
dir.create(outdir)
```


## Load Data
```{r}
#| label: load_data

# load stool metabolite data
met_data_path <- "../0_data/stoolMetaboliteData.xlsx"
metabo_data_stool <- readxl::read_xlsx(met_data_path)
colnames(metabo_data_stool) <- gsub("Indol", "Indole", colnames(metabo_data_stool))

# load serum metabolite data
met_data_path <- "../0_data/serumMetaboliteData.xlsx"
metabo_data_serum <- readxl::read_xlsx(met_data_path)
colnames(metabo_data_serum) <- gsub("Indol", "Indole", colnames(metabo_data_serum))

# load clinical data
clinic_data_path <- "../0_data/clinicData.xlsx"
clinic_data <- readxl::read_xlsx(clinic_data_path)
clinic_data <- clinic_data |> mutate(disease_cat = ifelse(is.na(disease_cat), "unknown", disease_cat))
```

## Function

Metabolite data was filtered for spurious patients. Then metabolites were removed with more than 90% missing values. Remaining missing values were imputed by half minimum and subsequently all values were log2 transformed. Batch correction was performed using ComBat with cancer type and antibiotics treatment as covariates being kept. 

```{r}
prefilter_metabolite_data <- function(metabolite_data,
                                      metadata,
                                      rm_samples,
                                      max_na_percent = 0.2,
                                      transform_data = TRUE,
                                      batch_correct_data = "ComBat",
                                      imputation_method = "zero") {
  
  # Remove data from spurious samples
  metabolite_data <- metabolite_data |> 
    filter(!(pat_id %in% rm_samples)) %>% 
    select(-sample_id)
  
  # Remove data not in metadata/metabolite_data
  shared_pat <- inner_join(metadata,
                           metabolite_data,
                           by = c("pat_id" = "pat_id")) |> 
    pull(pat_id)
  
  metadata <- metadata |> 
    filter(pat_id %in% shared_pat)
  metabolite_data <- metabolite_data |> 
    filter(pat_id %in% shared_pat)
  
  # filter metabolites for enough data 
  metabolite_data <- metabolite_data |> 
    pivot_longer(!c(pat_id)) |> 
    group_by(name) |> 
    mutate(select = (sum(is.na(value))/n()) < max_na_percent) |> 
    dplyr::filter(any(select)) |> 
    dplyr::select(-select) |> 
    pivot_wider()
  
  # NA imputation
  set.seed(18022025)
  rn <- as.vector(metabolite_data[,1])[[1]]
  df <- as.data.frame(metabolite_data[,-1])
  rownames(df) <- rn
  
  metabolite_data <- impute_NAs(data = df,
                                method = imputation_method) |> 
    as_tibble(rownames = "pat_id")
  
  # Transform Data
  if (transform_data) {
    metabolite_data[,-1] <- log2(metabolite_data[,-1])
  }
  
  # Batch correct Data
  if (batch_correct_data == "ComBat") {
    

    design <- model.matrix(~ metadata$disease_cat +
                             metadata$antibiotics_pretreatment_cat)
    
    
    df <- t(ComBat(dat = t(metabolite_data[,-1]),
                   batch = metadata$met_analysis_batch,
                   par.prior = F,
                   mod = design))
    
    rownames(df) <- rn
    
    metabolite_data <- as_tibble(df, rownames = "pat_id")
  } 
  
  return(metabolite_data)
}
```

## Prefilter Data
```{r}
# The response annotation of those patients has issues
spurious_samples <- c()

# spurious_samples <- c()

stool_data <- prefilter_metabolite_data(metabolite_data = metabo_data_stool,
                                        metadata = clinic_data,
                                        rm_samples = spurious_samples,
                                        max_na_percent = 0.9, 
                                        imputation_method = "HM") 

write_delim(stool_data,
            file = paste0(outdir, "/stoolMetaboliteData_preprocessed.tsv"),
            delim = "\t")

stool_metadata <- clinic_data |>
  filter(!(pat_id %in% spurious_samples),
         metabolomics_stool == 1) |> 
  mutate(disease_cat = ifelse(is.na(disease_cat), "unknown", disease_cat))

write_delim(stool_metadata,
            file = paste0(outdir, "/stoolClinicalData_preprocessed.tsv"),
            delim = "\t")


serum_data <- prefilter_metabolite_data(metabolite_data = metabo_data_serum,
                                        metadata = clinic_data,
                                        rm_samples = spurious_samples,
                                        max_na_percent = 0.9, 
                                        imputation_method = "HM") 

write_delim(serum_data,
            file = paste0(outdir, "/serumMetaboliteData_preprocessed.tsv"),
            delim = "\t")

serum_metadata <- clinic_data |>
  filter(!(pat_id %in% spurious_samples),
         metabolomics_serum == 1) |> 
  mutate(disease_cat = ifelse(is.na(disease_cat), "unknown", disease_cat))

write_delim(serum_metadata,
            file = paste0(outdir, "/serumClinicalData_preprocessed.tsv"),
            delim = "\t")
```

# Analysis

## Figure S1 - Influence of Abx on Survival and the Microbiome

### Set Paths and Load Data 
```{r}
figS1dir <- "../2_analysis/Figures/FigS1"
dir.create(figS1dir, recursive = T)
dir.create("../2_analysis/02_analysisData/")

# load clinical data
clinic_data_path <- "../0_data/clinicData.xlsx"
clinic_data_bac <- readxl::read_xlsx(clinic_data_path)
clinic_data_bac <- clinic_data_bac |>
  mutate(antibiotics_pretreatment_cat = ifelse(antibiotics_pretreatment_cat == "no Abx", "no Abx", "Abx"),
         antibiotics_pretreatment_cat = factor(antibiotics_pretreatment_cat, levels = c("no Abx", "Abx")))
saveRDS(clinic_data_bac, file = "../2_analysis/02_analysisData/clinic_data_bac.rds")

# load metagenomics data 
tax_data_path <- "../0_data/stoolMetagenomicsTaxData.xlsx"
tax_data <- readxl::read_xlsx(tax_data_path)
tax_data <- tax_data |> filter(pat_id %in% clinic_data_bac$pat_id)
```

### Abx Survival
```{r}

plt_survival <- function(data, surv_type = "PFS", color_vector) {
  
  if(surv_type == "PFS"){
    data <- data %>% 
      mutate(time = PFS_time, event = PFS_event) %>%
      select(time, event, antibiotics_pretreatment_cat)
  } else if(surv_type == "OS") {
    data <- data %>% 
      mutate(time = OS_time, event = OS_event) %>%
      select(time, event, antibiotics_pretreatment_cat)
  }

  # Fit survival curve
  fit <- survfit(Surv(time, event) ~ antibiotics_pretreatment_cat, data = data)
  LR_test <- survdiff(Surv(time, event) ~ antibiotics_pretreatment_cat, data = data)
  coxfit <- coxph(Surv(time, event) ~ antibiotics_pretreatment_cat, data = data)
  sumcox <- summary(coxfit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  time_cutoff = max(data$time)
  survs = summary(fit, times=time_cutoff)$surv
  labels = paste(round(survs*100), '%', sep='')
  
  surv <- ggsurvplot(fit,
                     data = data,
                     pval = FALSE,
                     conf.int = FALSE,
                     risk.table = TRUE,
                     ggtheme = theme_prism(), 
                     size = 1.5,
                     fontsize = 9/.pt)
    
  
  plt <- surv$plot + 
    scale_color_manual(values = color_vector, labels = c("no Abx", "Abx"))+
    labs(y = paste0("Probablity of ", surv_type," (%)"),
         x = "Days after CAR-T",
         subtitle = paste0(
           # "cox PH: p = ",
           # signif(sumcox$waldtest[3], 2),
           "p = ",
           signif(LR_test$pvalue, 2),
           ", HR = ",
           signif(exp(coxfit$coefficients), 2),
           " (",HR.confint.lower,"-",HR.confint.upper,")"),) +
    guides(colour = guide_legend(position = "inside")) +
    theme_prism(base_size = 9) +
    theme(legend.position.inside = c(0.8, 0.8),
          legend.key.height = unit(9, "pt")) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  risk_table_col <- color_vector
  risk_table_labs <- c("no Abx", "Abx")
  
 
  tp <- surv$table + 
    theme_prism(base_size = 9) +
    scale_color_manual(values = risk_table_col, label = risk_table_labs)+
    scale_fill_manual(values = risk_table_col, label = risk_table_labs) +
    scale_y_discrete(labels = rev(risk_table_labs)) +
    theme(axis.text.y = element_text(color = rev(risk_table_col))) +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
    rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    theme(plot.title = element_text(hjust = 0, size=9)) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')

  
  p2 <- ggarrange(plt, tp, ncol = 1, heights = c(3,1)) 
  
  p2 <- (plt / tp) + plot_layout(heights = c(5, 1))
  
  return(p2)
}

Abx_colors <- c("darkgrey", "darkorange")

Abx_PFS_plt <- plt_survival(clinic_data_bac, surv_type = "PFS", c("darkgrey", "darkorange"))
saveRDS(Abx_PFS_plt, file = paste0(figS1dir, "/01_Abx_PFS.rds"))
plot(Abx_PFS_plt)

Abx_OS_plt <- plt_survival(clinic_data_bac, surv_type = "OS",  c("darkgrey", "darkorange"))
saveRDS(Abx_OS_plt, file = paste0(figS1dir, "/02_Abx_OS.rds"))
plot(Abx_OS_plt)

```

### Alpha Diversity

#### Prepare Data
```{r}
# Create Phyloseq Object

# Extract Taxonomy
tax_mat <- tax_data |> 
  mutate(across(where(is.character), ~ gsub(".__", "", .))) |> 
  pivot_wider(id_cols = c(Kingdom, Phylum, Class, Order, Family, Genus, Species),
              names_from = pat_id, 
              values_from = abundance) |>
  dplyr::select(c(Kingdom, Phylum, Class, Order, Family, Genus, Species)) |> 
  as.matrix()
rownames(tax_mat) <- tax_mat[,7]

# Create Taxonomy table object
TAX <- tax_table(tax_mat)

# Extract OTU
otu_mat <- tax_data |> 
  pivot_wider(id_cols = c(Kingdom, Phylum, Class, Order, Family, Genus, Species),
              names_from = pat_id, 
              values_from = abundance) |>
  dplyr::select(-c(Kingdom, Phylum, Class, Order, Family, Genus, Species)) |> 
  as.matrix()
rownames(otu_mat) <- tax_mat[,7]

# Create OTU table object
OTU <- otu_table(otu_mat, taxa_are_rows = TRUE)

# sample_df <- clinic_data_bac |>
#   dplyr::select(response_CR_d180) |>
#   mutate(response_CR_d180 = as.factor(response_CR_d180)) |>
#   as.data.frame()



sample_df <- clinic_data_bac |>
  dplyr::select(response_CR_d180,
                PFS_event, OS_event,
                antibiotics_pretreatment_cat) |>
  mutate(response_CR_d180 = as.factor(response_CR_d180)) |>
  as.data.frame()

rownames(sample_df) <- clinic_data_bac$pat_id

SAMPLE <- sample_data(object = sample_df)

physeq <- phyloseq(OTU,TAX,SAMPLE)
saveRDS(physeq, file = "../2_analysis/02_analysisData/physeq.rds")

# Species found only in one sample:
otu_table(physeq) %>% as.data.frame() %>% as_tibble(rownames = "Species") %>% 
  pivot_longer(!Species) %>% 
  filter(value > 0) %>% 
  group_by(Species) %>% 
  summarise(n = n()) %>% 
  filter(n == 1)
```

#### Diversity Comparison Abx
```{r}
a_diversity <- microbiome::alpha(physeq) |> as_tibble(rownames = "pat_id") 

plt_alpha_diversity <- function(table, diversity_metric, yaxis_label, title_label) {
  table |> 
    dplyr::left_join(clinic_data_bac) |> 
    ggplot(aes(x = antibiotics_pretreatment_cat,
               y = !!sym(diversity_metric),
               color = antibiotics_pretreatment_cat)) +
    geom_quasirandom() +
    stat_boxplot(aes(ymin = after_stat(middle),
                     ymax = after_stat(middle)),
                 geom="errorbar",
                 linewidth=1,
                 width=.5, 
                 color = "black")  + 
    stat_compare_means(size = 8/.pt) +
    scale_color_manual( values = c( "darkgrey","darkorange")) +
    scale_fill_manual( values = c( "darkgrey","darkorange")) +
    labs(y = yaxis_label, 
         x = "",
         title = title_label) +
    theme_prism(base_size = 8) +
    theme(legend.position = "none")
}

plt_shannon <- plt_alpha_diversity(a_diversity,
                    diversity_metric = "diversity_shannon",
                    yaxis_label = "Shannon Index",
                    title_label = "diversity")

plt_dbp <- plt_alpha_diversity(a_diversity,
                    diversity_metric = "dominance_dbp",
                    yaxis_label = "Berger-Parker Index",
                    title_label = "dominance")

plt_rich <- plt_alpha_diversity(a_diversity,
                    diversity_metric = "observed",
                    yaxis_label = "observed richness",
                    title_label = "richness")

saveRDS(plt_rich, file = paste0(figS1dir, "/03_richness_abx.rds"))
saveRDS(plt_shannon, file = paste0(figS1dir, "/04_diversity_abx.rds"))
saveRDS(plt_dbp, file = paste0(figS1dir, "/05_dominance_abx.rds"))

ggarrange(plt_rich, plt_shannon, plt_dbp, nrow = 1)
```

#### Diversity and Survival

```{r}

plt_surv_adiversity <- function(data,
                              diversity_name = "chao1",
                              event_type = "PFS", 
                              color_palette = c("low" = "grey", "high" = "firebrick"),
                              man_cutoff = NA) {
  
  # Prepare survival data
  if(event_type == "PFS") {
    surv_dat <- data |> 
    # Combine metabolite data with clinical outcomes
      left_join(clinic_data_bac |> 
                  mutate(event = PFS_event,
                         time = as.numeric(PFS_time)) |>
                  dplyr::select(time, event, pat_id)) |> 
    # Convert to long format for analysis by metabolite
    pivot_longer(!c(time,
                    event,
                    pat_id)) |> 
    filter(name == diversity_name)
  } else if(event_type == "OS") {
    surv_dat <- data |> 
    # Combine metabolite data with clinical outcomes
    left_join(clinic_data |> 
                mutate(event = OS_event,
                       time = as.numeric(OS_time)) |>
                dplyr::select(time, event, pat_id)) |> 
    # Convert to long format for analysis by metabolite
    pivot_longer(!c(time,
                    event,
                    pat_id)) |> 
    filter(name == diversity_name)
  }
  
  # Get Cutpoint
  cutpoint_fit <- surv_cutpoint(surv_dat,
                                time = "time",
                                event = "event",
                                minprop = 0.1,  # Minimum proportion in each group
                                variables = "value")
  
  cutpoint_maxstat <- summary(cutpoint_fit)[1, 1]
  
  # Apply cutpoint
  if (is.na(man_cutoff)) {
    cutpoint <- cutpoint_maxstat
  } else if( !is.na(man_cutoff)) {
    cutpoint <- man_cutoff
  }
  
  surv_dat <- surv_dat |> 
      mutate(group = ifelse(value > cutpoint, "high", "low"))
  caption <- paste0("cutpoint:  <= ", signif(cutpoint, 3))
  
  
  # Extract metabolite name
  met_name <- surv_dat$name[1]
  
  # Create survival fit
  fit <- survminer::surv_fit(Surv(time, event) ~ group, data = surv_dat)
  LR_test <- survdiff(Surv(time, event) ~ group, data = surv_dat)
  coxfit <- coxph(Surv(time, event) ~ group, data = surv_dat)
  sumcox <- summary(coxfit)
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  time_cutoff = max(surv_dat$time)
  survs = summary(fit, times=time_cutoff)$surv
  
  # Generate survival plot
  surv <- ggsurvplot(fit,
                     pval = FALSE,          # Show p-value
                     conf.int = FALSE,     # Don't show confidence intervals
                     risk.table = TRUE,    # Show risk table
                     ggtheme = theme_prism(),
                     size = 1.5,
                     fontsize = 9/.pt)
  
  color_pal <- unname(color_palette)
  color_labs <- names(color_palette)
  
  # Customize plot appearance
  p1 <- surv$plot + 
    scale_color_manual(values = color_pal, labels = color_labs) +
    labs(y = paste0("Probability of ", event_type , " (%)"),
         x = "Days after CAR-T",
         # title = met_name,
         subtitle = paste0(
           # "cox PH: p = ",
           # signif(sumcox$waldtest[3], 2),
           "p = ",
           signif(LR_test$pvalue, 2),
           ", HR = ",
           signif(exp(coxfit$coefficients), 2),
           " (",HR.confint.lower,"-",HR.confint.upper,")"),
         caption = caption,
  ) +
  theme_prism(base_size = 9) +
  guides(colour = guide_legend(position = "inside")) +
  theme(legend.position.inside = c(0.8, 0.12),
        legend.key.height = unit(10, "pt"),
        plot.caption = element_text(vjust = 0)) +
  coord_cartesian(xlim = c(0, time_cutoff), 
                  clip = 'on')

tp <- surv$table + 
  theme_prism(base_size = 9) +
  scale_color_manual(values = color_pal, label = color_labs)+
  scale_fill_manual(values = color_pal, label = color_labs) +
  scale_y_discrete(labels = rev(color_labs)) +
  theme(axis.text.y = element_text(color =rev(color_pal))) +
  labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
  rremove("x.text") +
  rremove("x.ticks") +
  rremove("x.axis") +
  rremove("y.axis") +
  rremove("y.ticks") +
  theme(plot.title = element_text(hjust = 0, size=9)) +
  coord_cartesian(xlim = c(0, time_cutoff), 
                  clip = 'on')

  
  p2 <- ggarrange(p1, tp, ncol = 1, heights = c(3,1)) 
  
  p2 <- (free(p1, type = "label") / tp) + plot_layout(heights = c(5, 1))
  
  print(p2)
}




p1 <- plt_surv_adiversity(a_diversity,
                          "observed",
                          event_type = "PFS",
                          man_cutoff = quantile(a_diversity$observed)[2],
                          color_palette = c("high richness" = "hotpink", "low richness" = "black" ))

p2 <- plt_surv_adiversity(a_diversity,
                          "observed",
                          event_type = "OS",
                          man_cutoff = quantile(a_diversity$observed)[2],
                          color_palette = c("high richness" = "hotpink", "low richness" = "black" )) 

p3 <- plt_surv_adiversity(a_diversity,
                          "diversity_shannon",
                          event_type = "PFS",
                          man_cutoff = quantile(a_diversity$diversity_shannon)[2],
                          color_palette = c("high diversity" = "hotpink", "low diversity" = "black" ))

p4 <- plt_surv_adiversity(a_diversity,
                          "diversity_shannon",
                          event_type = "OS",
                          man_cutoff = quantile(a_diversity$diversity_shannon)[2],
                          color_palette = c("high diversity" = "hotpink", "low diversity" = "black" ))

p5 <- plt_surv_adiversity(a_diversity,
                          "dominance_dbp",
                          event_type = "PFS",
                          man_cutoff = quantile(a_diversity$dominance_dbp)[2],
                          color_palette = c("high dominance" = "hotpink", "low dominance" = "black" ))

p6 <- plt_surv_adiversity(a_diversity,
                          "dominance_dbp",
                          event_type = "OS",
                          man_cutoff = quantile(a_diversity$dominance_dbp)[2],
                          color_palette = c("high dominance" = "hotpink", "low dominance" = "black" ))
# Save Plots   
saveRDS(p1, file = paste0(figS1dir, "/06_richness_PFS.rds"))
saveRDS(p2, file = paste0(figS1dir, "/07_richness_OS.rds"))
saveRDS(p3, file = paste0(figS1dir, "/08_diversity_PFS.rds"))
saveRDS(p4, file = paste0(figS1dir, "/09_diversity_OS.rds"))
saveRDS(p5, file = paste0(figS1dir, "/10_dominance_PFS.rds"))
saveRDS(p6, file = paste0(figS1dir, "/11_dominance_OS.rds"))
```

## Figure S2 - Influence of Abx on Metabolite Production in Stool

### Set Paths and Load Data 
```{r}
figS2dir <- "../2_analysis/Figures/FigS2"
dir.create(figS2dir, recursive = T)

# load clinical data for stool data
clinic_data_path <- "../2_analysis/01_preprocessedData/stoolClinicalData_preprocessed.tsv"
clinic_data_stool <- read_delim(clinic_data_path, delim = "\t")
clinic_data_stool <- clinic_data_stool |>
  mutate(antibiotics_pretreatment_cat = ifelse(antibiotics_pretreatment_cat == "no Abx", "no Abx", "Abx"),
         antibiotics_pretreatment_cat = factor(antibiotics_pretreatment_cat, levels = c("no Abx", "Abx")))

# load stool metabolomics data 
metabolite_data_path <- "../2_analysis/01_preprocessedData/stoolMetaboliteData_preprocessed.tsv"
metabo_data_stool <- read_delim(metabolite_data_path, delim = "\t")
metabo_data_stool <- metabo_data_stool |> filter(pat_id %in% clinic_data_stool$pat_id)


# load clinical data for serum samples
clinic_data_path <- "../2_analysis/01_preprocessedData/serumClinicalData_preprocessed.tsv"
clinic_data_serum <- read_delim(clinic_data_path, delim = "\t")
clinic_data_serum <- clinic_data_serum |>
  mutate(antibiotics_pretreatment_cat = ifelse(antibiotics_pretreatment_cat == "no Abx", "no Abx", "Abx"),
         antibiotics_pretreatment_cat = factor(antibiotics_pretreatment_cat, levels = c("no Abx", "Abx")),
         response_CRS_grade = as.numeric(ifelse(response_CRS_grade == "2/HLH-Syndrom", 2, response_CRS_grade)))

# load serum metabolite data 
metabolite_data_path <- "../2_analysis/01_preprocessedData/serumMetaboliteData_preprocessed.tsv"
metabo_data_serum <- read_delim(metabolite_data_path, delim = "\t")
metabo_data_serum <- metabo_data_serum |> filter(pat_id %in% clinic_data_serum$pat_id)
```

### Functions
```{r}
plot_metabolite_violin <- function(metabolite_data,
                                   metadata,
                                   metabolites,
                                   conc_adjustment = 1,
                                   remove_outliers = "SD",
                                   units = "µmol/g") {

  # Prepare Data
  dat <- metabolite_data %>% 
    left_join(metadata %>% 
                select(pat_id,antibiotics_pretreatment_cat) %>% 
                mutate(antibiotics_pretreatment_cat = case_when(
                  pat_id %in% c("HY14", "BMU1", "UH_31072023") ~ "no Abx", # Gave stool sample before Abx treatment
                  TRUE ~ as.character(antibiotics_pretreatment_cat)
                )) %>% 
                mutate(antibiotics_pretreatment_cat = factor(antibiotics_pretreatment_cat, levels = c("no Abx", "Abx")))) %>% 
    select(antibiotics_pretreatment_cat,
           all_of(metabolites)) %>% 
    pivot_longer(!antibiotics_pretreatment_cat) |> 
    mutate(name = factor(name,
                         levels = metabolites))
  
  # Transform data back from log2
  dat_transformed <- dat |> 
    mutate(value = (2^value)*conc_adjustment)
  
  # Test significance
  sig <- dat_transformed |> 
    group_by(name) |>
    wilcox_test(value ~ antibiotics_pretreatment_cat, ref.group = "no Abx") |> 
    adjust_pvalue(method = "fdr") |> 
    add_significance() |> 
    mutate(p.adj = signif(p.adj, 2)) |> 
    add_x_position()
  
  clean_plt_dat <- dat_transformed |>
    group_by(name, antibiotics_pretreatment_cat) |> 
    mutate(median = median(value),
           mean = mean(value),
           IQR = stats::IQR(value),
           SD = sd(value),
           Q1 = quantile(value, probs = c(0.25), na.rm = FALSE),
           Q3 = quantile(value, probs = c(0.75), na.rm = FALSE),
           iqr_outlier_upper = Q3 + 1.5*IQR, 
           iqr_outlier_lower = Q1 - 1.5*IQR,
           sd_outlier_upper = mean + 2*SD, 
           sd_outlier_lower = mean - 2*SD,
    ) |>  
    ungroup() 
  
  if (remove_outliers == "SD") {
    clean_plt_dat <- clean_plt_dat %>% 
      filter(sd_outlier_lower <= value & value <= sd_outlier_upper)
  } else if (remove_outliers == "IQR") {
    clean_plt_dat <- clean_plt_dat %>% 
      filter(iqr_outlier_lower <= value & value <= iqr_outlier_upper)
  } else if (remove_outliers == "off") {
    clean_plt_dat <- clean_plt_dat 
  }
  
  
  sig_y_pos <- clean_plt_dat %>% 
    group_by(name) %>% 
    summarise(y.pos = max(value * 1.1)) %>% 
    pull(y.pos)
  
  n_metabolites <- length(unique(dat_transformed$name))
  
  clean_plt_dat   %>% 
    ggplot(aes(x = antibiotics_pretreatment_cat, y = value)) +
    geom_violin(aes(fill = antibiotics_pretreatment_cat), adjust = 2, scale = "count") +
    stat_boxplot(aes(ymin = median, ymax = median, y = median), geom="errorbar", linewidth=1, width=.5)  +
    facet_wrap(~ name, scale = "free", ncol = n_metabolites) +
    stat_pvalue_manual(sig,
                       hide.ns = F,
                       y.position = sig_y_pos,
                       tip.length = 0,
                       label = " p = {p}",
                       size = 7/.pt) +
    scale_fill_manual(values = c("no Abx" = "darkgrey", "Abx" = "darkorange")) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
    labs(x = "",
         y = paste0("Concentration [", units ,"]"),
         caption = "Wilcoxon test, median indicated, outliers not shown") +
    theme_prism(base_size = 9) +
    theme(plot.caption = element_text(size = 4.5, hjust = 0),
          strip.text.x = element_text(size = 9)) 
}
```

### Short Chain Fatty Acids
```{r}
metabolites <- c("Acetic acid",
                "Propionic acid",
                "Butyric acid",
                "Valeric acid",
                "Hexanoic acid")

plt_SCFA_stool <- plot_metabolite_violin(metabo_data_stool,
                                       clinic_data_stool, 
                                       metabolites,
                                       remove_outliers = "SD")

plot(plt_SCFA_stool)

saveRDS(plt_SCFA_stool, file = paste0(figS2dir, "/01_Abx_SCFA_stool.rds"))
```

### Branched Chain Fatty Acids
```{r}
metabolites <- c("Isobutyric acid",
                "2-Methyl butyric acid",
                "Isovaleric acid")

plt_BCFA_stool <- plot_metabolite_violin(metabo_data_stool,
                                       clinic_data_stool, 
                                       metabolites,
                                       remove_outliers = "SD")

plot(plt_BCFA_stool)

saveRDS(plt_BCFA_stool, file = paste0(figS2dir, "/02_Abx_BCFA_stool.rds"))
```

### Indoles
```{r}
metabolites <- c("Indole-3-acetic acid",
                 "Indole-3-carboxyaldehyd")

plt_Indoles_stool <- plot_metabolite_violin(metabo_data_stool,
                                       clinic_data_stool, 
                                       metabolites,
                                       remove_outliers = "SD")

plot(plt_Indoles_stool)

saveRDS(plt_Indoles_stool, file = paste0(figS2dir, "/03_Abx_Indoles_stool.rds"))
```
### Bile Acids
```{r}
metabolites <- c("Ursodeoxycholic acid",
                "Deoxycholic acid",
                "Lithocholic acid",
                "Chenodeoxycholic acid",
                "Cholic acid")


plt_BA_stool <- plot_metabolite_violin(metabo_data_stool,
                                       clinic_data_stool, 
                                       metabolites,
                                       remove_outliers = "SD",
                                       conc_adjustment = 0.001)

plot(plt_BA_stool)

saveRDS(plt_BA_stool, file = paste0(figS2dir, "/04_Abx_BA_stool.rds"))
```

## Figure 1 - Single Species are not Predictive for CAR-T Outcomes

### Set Paths and Load Data 
```{r}
fig1dir <- "../2_analysis/Figures/Fig1"
dir.create(fig1dir, recursive = T)

# load metagenomics data
pw_data_path <- "../0_data/stoolMetagenomicsPWData.xlsx"
pw_data <- readxl::read_xlsx(pw_data_path)
pw_data <- pw_data |> filter(pat_id %in% clinic_data_bac$pat_id)
```

### Outcome Associations

#### LEfSe 
```{r}
run_lefse_analysis <- function(physeq, taxa_level = "Species", outcome_cols) {
  
  # Agglomerate to Taxa level
  physeq <- tax_glom(physeq, taxrank = taxa_level)
  physeq_rel  = transform_sample_counts(physeq, function(x) x / sum(x) )
  counts <- unclass(otu_table(physeq_rel))
  rownames(counts) <- tax_table(physeq_rel)[, taxa_level] %>% as.vector()
  coldata <- as(sample_data(physeq_rel), "data.frame")
  
  ## Create a SummarizedExperiment object
  se <- SummarizedExperiment(assays = list(counts = counts), colData = coldata)
  
  ## Preprocess Data for Lefse
  tn <- get_terminal_nodes(rownames(se))
  se_tn <- se[tn,]
  
  se_tn_ra <- relativeAb(se_tn)
  
  ## Adjust Labels
  colData(se_tn_ra)$OS_event <- factor(colData(se_tn_ra)$OS_event, levels = c(1,0), labels = c("no survival", "survival"))
  colData(se_tn_ra)$PFS_event <- factor(colData(se_tn_ra)$PFS_event, levels = c(1,0), labels = c("no PFS", "PFS"))
  colData(se_tn_ra)$response_CR_d180 <- fct_recode(colData(se_tn_ra)$response_CR_d180, "no CR d180" = "0", "CR d180" = "1")
  
  ## Plot
  res <- lefser(se_tn_ra, classCol = outcome_cols)  
  p1 <- lefserPlot(res)
  res_summary <- res %>% 
    as_tibble() %>% 
    mutate(Taxa = features, 
           direction = ifelse(scores > 0, "positive", "negative"),
           analysis = "LEFSE",
           outcome = outcome_cols,
           Taxa_level = taxa_level) %>% 
    select(Taxa, direction, analysis, outcome, Taxa_level)
  
  return(list(plot = p1, res_summary = res_summary))
}

# Species
res_lefse_spec_os <- run_lefse_analysis(physeq, taxa_level = "Species", outcome_cols = "OS_event")
res_lefse_spec_pfs <- run_lefse_analysis(physeq, taxa_level = "Species", outcome_cols = "PFS_event")
res_lefse_spec_cr <- run_lefse_analysis(physeq, taxa_level = "Species", outcome_cols = "response_CR_d180")
# Genus
res_lefse_gen_os <- run_lefse_analysis(physeq, taxa_level = "Genus", outcome_cols = "OS_event")
res_lefse_gen_pfs <- run_lefse_analysis(physeq, taxa_level = "Genus", outcome_cols = "PFS_event")
res_lefse_gen_cr <- run_lefse_analysis(physeq, taxa_level = "Genus", outcome_cols = "response_CR_d180")
```

#### Univariate Cox PH model

```{r}

run_cox_univariate <- function(tax_data, clinic_data, 
                               abundance_col = "abundance",
                               taxa_level = "Species", 
                               patient_col = "pat_id",
                               time_col = "PFS_time",
                               event_col = "PFS_event",
                               prevalence_thresh = 0.2,
                               binarize = FALSE,
                               plot_title = NULL,
                               plot_subtitle = NULL){
  
  if(binarize){
    abundance_label <- "prevalence"
  } else {
    abundance_label <- "abundance"
  }
  
  n_samples <- length(unique(tax_data[[patient_col]]))
  
  model_dat <- tax_data %>%
    group_by(across(all_of(c(taxa_level, patient_col)))) %>%
    summarise(abundance = sum(.data[[abundance_col]]), .groups = "drop") %>%
    group_by(across(all_of(taxa_level))) %>%
    mutate(prevalence = sum(abundance > 0)) %>%
    filter(prevalence > prevalence_thresh * n_samples) %>%
    { 
      if(binarize){
        mutate(., abundance = ifelse(abundance > 0, 1, 0))
      } else {
        .
      }
    } %>%
    ungroup() %>%
    select(all_of(patient_col), all_of(taxa_level), abundance) %>%
    pivot_wider(names_from = !!sym(taxa_level), values_from = abundance, values_fill = 0)
  
  surv_dat <- model_dat %>%
    left_join(
      clinic_data %>%
        select(all_of(patient_col), all_of(time_col), all_of(event_col)) %>%
        dplyr::rename(time = !!sym(time_col),
               event = !!sym(event_col)),
      by = patient_col
    )
  
  # Covariates are all columns except patient_col, time, event
  covariates <- setdiff(colnames(surv_dat), c(patient_col, "time", "event"))
  univ_formulas <- lapply(covariates,
                          function(x) as.formula(paste("Surv(time, event) ~", x)))
  names(univ_formulas) <- covariates
  
  univ_models <- lapply(univ_formulas,
                        function(f) coxph(f, data = surv_dat))
  
  univ_results <- lapply(univ_models, function(x){
    s <- summary(x)
    p.value <- signif(s$wald["pvalue"], digits = 2)
    wald.test <- signif(s$wald["test"], digits = 2)
    beta <- signif(s$coef[1], digits=2)
    HR <- signif(s$coef[2], digits=2)
    HR.confint.lower <- signif(s$conf.int[,"lower .95"], 2)
    HR.confint.upper <- signif(s$conf.int[,"upper .95"], 2)
    HRtotal <- paste0(HR, " (", HR.confint.lower, "-", HR.confint.upper, ")")
    res <- c(beta, HR, HR.confint.lower, HR.confint.upper, HRtotal, wald.test, p.value)
    names(res) <- c("beta", "HR", "conf.lower", "conf.upper", "HR (95% CI for HR)", "wald.test", "p.value")
    return(res)
  })
  
  res <- as_tibble(t(as.data.frame(univ_results, check.names = FALSE)), rownames = "Taxa") %>%
    mutate(p.value = as.numeric(p.value),
           HR = as.numeric(HR)) %>%
    arrange(p.value)
  
  res_summary <- res %>% 
    filter(p.value < 0.05) %>% 
    mutate(Taxa = gsub("^[a-z]__", "", Taxa), 
           direction = ifelse(HR > 1, "negative", "positive"),
           analysis = paste0("univ. cox reg ", abundance_label),
           outcome = event_col,
           Taxa_level = taxa_level) %>% 
    select(Taxa, direction, analysis, outcome, Taxa_level)
  
  if(is.null(plot_title)){
    plot_title <- ifelse(binarize, 
                         "Univariate Cox Regression PFS with Prevalence", 
                         "Univariate Cox Regression PFS with Abundance")
  }
  
  if(is.null(plot_subtitle)) {
    plot_subtitle <- paste0("Prevalence of at least ", prevalence_thresh * 100, "% required")
  }
  
  p <- res %>%
    filter(p.value < 0.05 & HR < 100) %>%
    mutate(col = ifelse(HR >= 1, "HR > 1", "HR < 1"),
           Taxa = factor(Taxa, levels = Taxa[order(HR)])) %>%
    ggplot(aes(x = Taxa, y = HR - 1, fill = col)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = function(y) y + 1) +
    coord_flip() +
    scale_fill_manual(values = rev(c("HR > 1" = "#E57A77", "HR < 1" = "#7CA1CC"))) +
    labs(y = "HR", x = NULL,
         title = plot_title,
         subtitle = plot_subtitle) +
    theme_prism(base_size = 10)
  
  return(list(plot = p, res_summary = res_summary))
}


res_cox_abund_spec_pfs <- run_cox_univariate(tax_data, clinic_data_bac, binarize = FALSE,
                                             time_col = "PFS_time", event_col = "PFS_event", taxa_level = "Species",
                                             plot_title = "Univariate Cox Regression PFS with Abundance")

res_cox_prev_spec_pfs <- run_cox_univariate(tax_data, clinic_data_bac, binarize = TRUE,
                                            time_col = "PFS_time", event_col = "PFS_event", taxa_level = "Species",
                                            plot_title = "Univariate Cox Regression PFS with Prevalence")

res_cox_abund_spec_os <- run_cox_univariate(tax_data, clinic_data_bac, binarize = FALSE,
                                            time_col = "OS_time", event_col = "OS_event", taxa_level = "Species",
                                            plot_title = "Univariate Cox Regression OS with Abundance")

res_cox_prev_spec_os <- run_cox_univariate(tax_data, clinic_data_bac, binarize = TRUE,
                                           time_col = "OS_time", event_col = "OS_event", taxa_level = "Species",
                                           plot_title = "Univariate Cox Regression OS with Prevalence")

res_cox_abund_gen_pfs <- run_cox_univariate(tax_data, clinic_data_bac, binarize = FALSE,
                                             time_col = "PFS_time", event_col = "PFS_event", taxa_level = "Genus",
                                             plot_title = "Univariate Cox Regression PFS with Abundance")

res_cox_prev_gen_pfs <- run_cox_univariate(tax_data, clinic_data_bac, binarize = TRUE,
                                            time_col = "PFS_time", event_col = "PFS_event", taxa_level = "Genus",
                                            plot_title = "Univariate Cox Regression PFS with Prevalence")

res_cox_abund_gen_os <- run_cox_univariate(tax_data, clinic_data_bac, binarize = FALSE,
                                            time_col = "OS_time", event_col = "OS_event", taxa_level = "Genus",
                                            plot_title = "Univariate Cox Regression OS with Abundance")

res_cox_prev_gen_os <- run_cox_univariate(tax_data, clinic_data_bac, binarize = TRUE,
                                           time_col = "OS_time", event_col = "OS_event", taxa_level = "Genus",
                                           plot_title = "Univariate Cox Regression OS with Prevalence")
```


#### LASSO-penalized Cox PH model

```{r}

run_cox_lasso <- function(tax_data, clinic_data, 
                          abundance_col = "abundance",
                          taxa_level = "Species", 
                          patient_col = "pat_id",
                          time_col = "PFS_time",
                          event_col = "PFS_event",
                          prevalence_thresh = 0.2,
                          binarize = FALSE,
                          plot_title = NULL,
                          plot_subtitle = NULL,
                          nfolds = 10,
                          glmnet_alpha = 1,
                          seed = 123){
  set.seed(seed)
  
  if(binarize){
    abundance_label <- "prevalence"
  } else {
    abundance_label <- "abundance"
  }
  
  n_samples <- length(unique(tax_data[[patient_col]]))
  
  # Prepare data similar to univariate function
  model_dat <- tax_data %>%
    group_by(across(all_of(c(taxa_level, patient_col)))) %>%
    summarise(abundance = sum(.data[[abundance_col]]), .groups = "drop") %>%
    group_by(across(all_of(taxa_level))) %>%
    mutate(prevalence = sum(abundance > 0)) %>%
    filter(prevalence > prevalence_thresh * n_samples) %>%
    { 
      if(binarize){
        mutate(., abundance = ifelse(abundance > 0, 1, 0))
      } else {
        .
      }
    } %>%
    ungroup() %>%
    select(all_of(patient_col), all_of(taxa_level), abundance) %>%
    pivot_wider(names_from = !!sym(taxa_level), values_from = abundance, values_fill = 0)
  
  surv_dat <- model_dat %>%
    left_join(
      clinic_data %>%
        select(all_of(patient_col), all_of(time_col), all_of(event_col)) %>%
        dplyr::rename(time = !!sym(time_col),
                      event = !!sym(event_col)),
      by = patient_col
    ) %>%
    filter(time > 0) %>% 
    drop_na(time, event) # ensure no missing survival info
  
  # Prepare matrix for glmnet
  x <- as.matrix(surv_dat %>% select(-all_of(c(patient_col, "time", "event"))))
  y <- surv_dat %>% select(time, event)
  surv_obj <- with(y, Surv(time, event))
  
  # Run cross-validated LASSO Cox model
  cvfit <- cv.glmnet(x, surv_obj, family = "cox", nfolds = nfolds, alpha = glmnet_alpha)
  
  # Coefficients at lambda.min
  coef_min <- coef(cvfit, s = "lambda.min")
  selected_idx <- which(coef_min != 0)
  
  if(length(selected_idx) == 0){
    warning("No features selected at lambda.min.")
    return(NULL)
  }
  
  selected_coefs <- data.frame(
    Taxa = rownames(coef_min)[selected_idx],
    coef = as.numeric(coef_min[selected_idx])
  ) %>%
    arrange(coef)
  
  if(is.null(plot_title)){
    plot_title <- ifelse(binarize, 
                         "LASSO Cox Regression PFS with Prevalence", 
                         "LASSO Cox Regression PFS with Abundance")
  }
  
  if(is.null(plot_subtitle)) {
    plot_subtitle <- paste0("Prevalence of at least ", prevalence_thresh * 100, "% required\nSelected features at lambda.min")
  }
  
  p <- ggplot(selected_coefs, aes(x = reorder(Taxa, coef), y = coef, fill = ifelse(coef < 0, "Positive", "Negative"))) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c("Positive" = "#7CA1CC" , "Negative" = "#E57A77")) +
    labs(x = NULL, y = "LASSO Coefficient",
         title = plot_title,
         subtitle = plot_subtitle) +
    theme_prism(base_size = 10)
  
  
  res_summary <- selected_coefs %>% 
    mutate(Taxa = gsub("^[a-z]__", "", Taxa), 
           direction = ifelse(coef > 0, "negative", "positive"),
           analysis = paste0("LASSO cox reg ", abundance_label),
           outcome = event_col,
           Taxa_level = taxa_level) %>% 
    select(Taxa, direction, analysis, outcome, Taxa_level)
    
  return(list(cvfit = cvfit, selected_coefs = selected_coefs, plot = p, res_summary = res_summary))
}

res_lasso_abund_spec_pfs <- run_cox_lasso(tax_data, clinic_data_bac, binarize = FALSE, prevalence_thresh = 0)
res_lasso_prev_spec_pfs <- run_cox_lasso(tax_data, clinic_data_bac, binarize = TRUE, prevalence_thresh = 0)
res_lasso_abund_spec_os <- run_cox_lasso(tax_data, clinic_data_bac, binarize = FALSE, prevalence_thresh = 0, time_col = "OS_time", event_col = "OS_event")
res_lasso_prev_spec_os <- run_cox_lasso(tax_data, clinic_data_bac, binarize = TRUE, prevalence_thresh = 0, time_col = "OS_time", event_col = "OS_event")

res_lasso_abund_gen_pfs <- run_cox_lasso(tax_data, clinic_data_bac, binarize = FALSE, prevalence_thresh = 0, taxa_level = "Genus")
res_lasso_prev_gen_pfs <- run_cox_lasso(tax_data, clinic_data_bac, binarize = TRUE, prevalence_thresh = 0, taxa_level = "Genus")
res_lasso_abund_gen_os <- run_cox_lasso(tax_data, clinic_data_bac, binarize = FALSE, prevalence_thresh = 0, time_col = "OS_time", event_col = "OS_event", taxa_level = "Genus")
res_lasso_prev_gen_os <- run_cox_lasso(tax_data, clinic_data_bac, binarize = TRUE, prevalence_thresh = 0, time_col = "OS_time", event_col = "OS_event", taxa_level = "Genus")
```

#### LinDA
```{r}

run_LinDA_analysis <- function(physeq, taxa_level = "Species", outcome_cols) {
  pseq <- physeq
  phyloseq::sample_data(pseq)$response_CR_d180 <- fct_recode(phyloseq::sample_data(physeq)$response_CR_d180, "no CR" = "0", "CR" = "1")
  phyloseq::sample_data(pseq)$OS_event <- factor(phyloseq::sample_data(physeq)$OS_event, levels = c(1,0), labels = c("no OS", "OS"))
  phyloseq::sample_data(pseq)$PFS_event <- factor(phyloseq::sample_data(physeq)$PFS_event, levels = c(1,0), labels = c("no PFS", "PFS"))
  
  mStat_obj <- mStat_convert_phyloseq_to_data_obj(pseq)
  res.list <- generate_taxa_test_single(data.obj = mStat_obj,
                                        group.var = outcome_cols,
                                        feature.dat.type = "proportion",
                                        feature.level = taxa_level,
                                        prev.filter = 0.1)
  res_summary <- res.list[[1]][[1]] %>% 
    filter(Adjusted.P.Value < 0.05) %>% 
    mutate(Taxa = Variable,
           direction = ifelse(Coefficient < 0, "positive", "negative"),
           analysis = "LinDA",
           outcome = outcome_cols,
           Taxa_level = taxa_level) %>% 
    mutate(direction = as.character(direction)) %>% 
    select(Taxa, direction, analysis, outcome, Taxa_level)
  
  
  volcano_plot <- generate_taxa_volcano_single(
    data.obj = mStat_obj,
    group.var = outcome_cols,
    test.list = res.list,
    feature.sig.level = 0.1,
    feature.mt.method = "fdr")
  
  return(list(plot = volcano_plot[[1]][[1]], res_summary = res_summary))
}


res_linda_spec_cr <- run_LinDA_analysis(physeq, "Species", "response_CR_d180")
res_linda_spec_os <- run_LinDA_analysis(physeq, "Species", "OS_event")
res_linda_spec_pfs <- run_LinDA_analysis(physeq, "Species", "PFS_event")
res_linda_gen_cr <- run_LinDA_analysis(physeq, "Genus", "response_CR_d180")
res_linda_gen_os <- run_LinDA_analysis(physeq, "Genus", "OS_event")
res_linda_gen_pfs <- run_LinDA_analysis(physeq, "Genus", "PFS_event")
```

#### MaAsLin3

```{r}
# Maaslin -----------------------------------------------------------------
run_maaslin_for_taxonomic_level <- function(physeq, tax_rank, outcome = "OS_event", 
                                            output_dir_base) {
  # Define output directory path
  out_dir <- file.path(output_dir_base, paste0("maaslin3_", outcome, "_", tax_rank))
  
  # Skip if output directory exists (assumes successful previous run)
  if (dir.exists(out_dir)) {
    message(paste("Skipping", outcome, tax_rank, "- output directory already exists:", out_dir))
    return(NULL)
  }
  
  # Aggregate to required taxonomic level
  physeq_tax <- tax_glom(physeq, tax_rank)
  
  # Extract taxon names
  taxa_names <- tax_table(physeq_tax) %>% as.data.frame() %>% pull(!!sym(tax_rank))
  
  # Prepare taxa table (OTU abundances)
  taxa_table <- otu_table(physeq_tax) %>% as.matrix() %>% as.data.frame() %>% t()
  colnames(taxa_table) <- taxa_names
  
  # Prepare metadata
  metadata <- sample_data(physeq_tax) %>% as.matrix() %>% as.data.frame()
  metadata$Sample <- rownames(metadata)
  metadata <- metadata %>%
    select(Sample, response_CR_d180, OS_event, PFS_event, antibiotics_pretreatment_cat)
  
  metadata$response_CR_d180 <- factor(metadata$response_CR_d180, levels = c(0,1), labels = c("no CR d180", "CR d180"))
  metadata$PFS_event <- factor(metadata$PFS_event, levels = c(1,0), labels = c("no PFS", "PFS"))
  metadata$OS_event <- factor(metadata$OS_event, levels = c(1,0), labels = c("no PFS", "PFS"))
  
  formula_str <- paste("~", outcome)
  out_dir <- file.path(output_dir_base, paste0("maaslin3_", outcome, "_", tax_rank))
  maaslin3_fit_out <- maaslin3(
    input_data = taxa_table,
    input_metadata = metadata,
    output = out_dir,
    formula = formula_str,
    cores = 2
  )
}

dir.create("../2_analysis/Differential_Abundance/")
run_maaslin_for_taxonomic_level(physeq, "Species", outcome = "response_CR_d180", output_dir_base = "../2_analysis/Differential_Abundance/")
run_maaslin_for_taxonomic_level(physeq, "Species", outcome = "PFS_event", output_dir_base = "../2_analysis/Differential_Abundance/")
run_maaslin_for_taxonomic_level(physeq, "Species", outcome = "OS_event", output_dir_base = "../2_analysis/Differential_Abundance/")

run_maaslin_for_taxonomic_level(physeq, "Genus", outcome = "response_CR_d180", output_dir_base = "../2_analysis/Differential_Abundance/")
run_maaslin_for_taxonomic_level(physeq, "Genus", outcome = "PFS_event", output_dir_base = "../2_analysis/Differential_Abundance/")
run_maaslin_for_taxonomic_level(physeq, "Genus", outcome = "OS_event", output_dir_base = "../2_analysis/Differential_Abundance/")

```

#### Upset Plot
```{r}
aggregated_results <- bind_rows(res_lefse_spec_os$res_summary,
                                res_lefse_spec_pfs$res_summary,
                                res_lefse_spec_cr$res_summary,
                                res_lefse_gen_os$res_summary,
                                res_lefse_gen_pfs$res_summary,
                                res_lefse_gen_cr$res_summary,
                                res_cox_abund_spec_pfs$res_summary,
                                res_cox_prev_spec_pfs$res_summary,
                                res_cox_abund_spec_os$res_summary,
                                res_cox_prev_spec_os$res_summary,
                                res_cox_abund_gen_pfs$res_summary,
                                res_cox_prev_gen_pfs$res_summary,
                                res_cox_abund_gen_os$res_summary,
                                res_cox_prev_gen_os$res_summary,
                                res_lasso_abund_spec_pfs$res_summary,
                                res_lasso_prev_spec_pfs$res_summary,
                                res_lasso_abund_spec_os$res_summary,
                                res_lasso_prev_spec_os$res_summary,
                                res_lasso_abund_gen_pfs$res_summary,
                                res_lasso_prev_gen_pfs$res_summary,
                                res_lasso_abund_gen_os$res_summary,
                                res_lasso_prev_gen_os$res_summary,
                                res_linda_spec_cr$res_summary,
                                res_linda_spec_os$res_summary,
                                res_linda_spec_pfs$res_summary,
                                res_linda_gen_cr$res_summary,
                                res_linda_gen_os$res_summary,
                                res_linda_gen_pfs$res_summary,
                                ) 


aggregated_results %>% 
  filter(analysis == "LinDA")

plt <- aggregated_results %>% 
  mutate(label = paste(analysis, outcome)) %>% 
  select(-analysis, -outcome, -Taxa_level) %>% 
  group_by(Taxa) %>% 
  summarize(label = list(label)) %>%  
  ggplot(aes(x=label)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1, size = 7/.pt) +
  scale_x_upset(n_intersections = 30) +
  theme_prism(base_size = 9) +
  labs(x = NULL, y = "Species and Genera [n]",
       subtitle = "Significant Taxa Found for different tools and outcome",
       caption = "LinDA OS_event, LinDA PFS_event, Maaslin3 OS_event, Maaslin3 PFS_event and Maaslin3 response_CR_d180 had no significant associations")+
  theme_combmatrix(combmatrix.label.text = element_text(size=5),
                   combmatrix.label.extra_spacing = 0.5,
                   combmatrix.label.total_extra_spacing =  unit(0.5, "pt"),
                   combmatrix.panel.point.size = 1.5,
                   combmatrix.panel.line.size = .4,
                   combmatrix.label.height = unit(70, "pt"),
                   combmatrix.panel.margin = unit(c(0, 0), "pt"))


plt
saveRDS(plt, file = paste0(fig1dir, "/01_UpsetPlot.rds"))
```


### Previously Described Species

#### Load Data

```{r}
described_data_path <- "../0_data/Described_Bacteria_v2.xlsx" 
pub_data <- readxl::read_xlsx(described_data_path)
```

#### Histogram
```{r}
# Create a function to perform the Cox PH analysis for a given taxonomic level
run_coxph_analysis <- function(tax_data, pub_data, clinic_data, tax_level, predictor_var = "abundance") {
  tax_data %>% 
    # Filter for taxa in publication data
    filter(!!sym(tax_level) %in% pub_data[[tax_level]]) %>%
    # Summarise by selected taxa level
    group_by(pat_id, !!sym(tax_level)) %>% 
    summarise(abundance = sum(abundance)) %>% 
    # Create presence variable and calculate prevalence
    mutate(presence = abundance > 0) %>% 
    group_by(!!sym(tax_level)) %>% 
    mutate(prevalence = sum(abundance > 0.01)) %>% 
    # Join with clinical data
    dplyr::left_join(clinic_data %>% select(pat_id, PFS_event, PFS_time, OS_time, OS_event)) %>% 
    group_by(!!sym(tax_level)) %>% 
    nest() %>% 
    # Run Cox PH models for both OS and PFS
    mutate(
      OS_coxPH = map(data, function(x){
        coxfit <- coxph(Surv(OS_time, OS_event) ~ !!sym(predictor_var), data = x)
        return(broom::tidy(coxfit))
      }),
      PFS_coxPH = map(data, function(x){
        coxfit <- coxph(Surv(PFS_time, PFS_event) ~ !!sym(predictor_var), data = x)
        return(broom::tidy(coxfit))
      })) %>% 
    unnest(PFS_coxPH)
}

# Apply the function to Species level using abundance as predictor
Species_coxph <- run_coxph_analysis(
  tax_data = tax_data,
  pub_data = pub_data,
  clinic_data = clinic_data,
  tax_level = "Species",
  predictor_var = "presence"
)

# Apply the function to Genus level using presence as predictor
Genus_coxph <- run_coxph_analysis(
  tax_data = tax_data,
  pub_data = pub_data,
  clinic_data = clinic_data,
  tax_level = "Genus",
  predictor_var = "presence"
)

# Combine results
results_described_species <- bind_rows(Species_coxph, Genus_coxph)


# Plot Histogram
plt_cox_hist <- results_described_species %>% 
  ggplot(aes(x = p.value)) +
  geom_histogram(binwidth = 0.1, color = "black") +
  geom_hline(yintercept = nrow(results_described_species)/10, color = "red") +
  labs(title = "Significance of univariate coxPH models",
       subtitle = "previously described Species and Genera",
       x = "p value",
       y = "n") +
  theme_prism(base_size = 9, border = T) +
  scale_y_continuous(breaks = seq(0, 13, by = 5), minor_breaks = seq(0,13, by = 1)) 

plt_cox_hist

saveRDS(plt_cox_hist, file = paste0(fig1dir, "/02_coxPH_hist.rds"))
```

### Analyze Selected Bacteria
#### Functions

```{r}
# Functions ---------------------------------------------------------------

### Wrapper for Metabolite plot
get_met_plot <- function(bac_dat,
                         met_dat, 
                         selected_metabolite,
                         units = "µmol/g",
                         low_col = "grey80",
                         high_col = "grey30") {
  require(colorspace)
  
  # Set Colors
  fixed_met_colors <- c("Valeric acid" = "#BE97DE",
                        "Isovaleric acid" = "#CCE4FD", 
                        "Indole-3-acetic acid" = "#B37F79",
                        "Indole-3-carboxyaldehyd" = "#FCC384" )
  
  if(selected_metabolite %in% names(fixed_met_colors)) {
    base_color = fixed_met_colors[[selected_metabolite]]
    low_col = base_color 
    high_col = darken(base_color, 0.3)
  }

  # Filter for specific metabolite
  met_dat <- met_dat |> 
    pivot_longer(!pat_id,
                 names_to = "metabolite",
                 values_to = "concentration") |> 
    dplyr::filter(metabolite == selected_metabolite)
  
  # Join data
  bac_met_dat <- bac_dat |> 
    select(pat_id, name, value, group) |> 
    dplyr::inner_join(met_dat) |> 
    # Convedrt log2 values back
    mutate(concentration = 2^concentration)
  
  # Plot
  sig <- bac_met_dat |> 
    wilcox_test(concentration ~ group) |> 
    add_x_position()
  
  clean_plt_dat <-  bac_met_dat |>
    group_by(group) |> 
    mutate(median = median(concentration),
           IQR = stats::IQR(concentration),
           Q1 = quantile(concentration, probs = c(0.25), na.rm = FALSE),
           Q3 = quantile(concentration, probs = c(0.75), na.rm = FALSE),
           outlier_upper = Q3 + 1.5*IQR, 
           outlier_lower = Q1 - 1.5*IQR) |>  
    filter(outlier_lower <= concentration & concentration <= outlier_upper)
  
  sig_y_pos <- max(clean_plt_dat$concentration) * 1.1
  n_groups <- bac_met_dat |>
    mutate(group = as.numeric(group)) %>% 
    dplyr::count(group)
  n_low <- n_groups |> filter(group == "1") |> pull(n) # 1 or 0 factor
  n_high <- n_groups |> filter(group == "2") |> pull(n)
  
  lab_low <- paste0(levels(bac_met_dat$group)[1], " \n n = ", n_low)
  lab_high <- paste0(levels(bac_met_dat$group)[2], " \n n = ", n_high)
  
  clean_plt_dat |> 
    ggplot(aes(x = group, y = concentration)) +
    geom_violin(aes(fill = group)) +
    stat_boxplot(aes(ymin = median, ymax = median, y = median), geom="errorbar", linewidth=1, width=.5)  + 
    stat_pvalue_manual(sig,
                       hide.ns = F,
                       y.position = sig_y_pos,
                       tip.length = 0,
                       label = " p = {p}",
                       size = 3) +
    scale_fill_manual(values = c(low_col,high_col)) +
    scale_x_discrete(labels = c(lab_low,
                                lab_high)) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
    labs(x = bac_met_dat$name[1],
         y = paste0(selected_metabolite,
                    " [",units,"]"),
         caption = "Wilcoxon test, median indicated, outliers not shown") +
    theme_prism(base_size = 9) +
    theme(plot.caption = element_text(size = 4.5, hjust = 0))
}

### Wrapper for KM plot
create_KM_plot <- function(survival_dat,
                           event_type,
                           low_col = "grey80",
                           high_col = "grey30") {
  
  # Create survival fit
  fit <- survminer::surv_fit(Surv(time, event) ~ group, data = survival_dat)
  LR_test <- survdiff(Surv(as.numeric(time), event) ~ group, data = survival_dat)
  coxfit <- coxph(Surv(time, event) ~ group, data = survival_dat)
  sumcox <- summary(coxfit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  caption <- paste0("cutpoint:  <= ", survival_dat$cutpoint)
  
  time_cutoff = max(survival_dat$time)
  survs = summary(fit, times=time_cutoff)$surv
  
  # Generate survival plot
  surv <- ggsurvplot(fit,
                     pval = FALSE,          # Show p-value
                     conf.int = FALSE,     # Don't show confidence intervals
                     risk.table = TRUE,    # Show risk table
                     size = 1.5,
                     fontsize = 9/.pt)
  
  # Customize plot appearance
  p1 <- surv$plot + 
    scale_color_manual(values = c(low_col, high_col),
                       labels = c(levels(survival_dat$group)[1], 
                                  levels(survival_dat$group)[2])) +
    labs(y = paste0("Probability of ", event_type , " (%)"),
         x = "Days after CAR-T",
         title = survival_dat$name[1],
         subtitle = paste0(
           # "cox PH: p = ",
           # signif(sumcox$waldtest[3], 2),
           "p = ",
           signif(LR_test$pvalue, 2),
           ", HR = ",
           signif(exp(coxfit$coefficients), 2),
           " (",HR.confint.lower,"-",HR.confint.upper,")"),
         caption = caption
    ) +
    theme_prism(base_size = 9) +
    guides(colour = guide_legend(position = "inside")) +
    theme(legend.position.inside = c(0.8, 0.12),
          legend.key.height = unit(10, "pt")) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  risk_table_col <- c(low_col, high_col)
  risk_table_labs <- c(levels(survival_dat$group)[1], 
                                  levels(survival_dat$group)[2])
  
 
  tp <- surv$table + 
    theme_prism(base_size = 9) +
    scale_color_manual(values = risk_table_col, label = risk_table_labs)+
    scale_fill_manual(values = risk_table_col, label = risk_table_labs) +
    scale_y_discrete(labels = rev(risk_table_labs)) +
    theme(axis.text.y = element_text(color =rev(risk_table_col))) +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
    rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    theme(plot.title = element_text(hjust = 0, size=9)) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')

  
  p2 <- ggarrange(p1, tp, ncol = 1, heights = c(3,1)) 
  
  p2 <- (p1 / tp) + plot_layout(heights = c(3, 1))
  
  
  return(p2)
}

get_OS_plot <- function(dat, high_col, low_col) {
  surv_dat <- dat |> 
    mutate(time = OS_time,
           event = OS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "OS", high_col = high_col, low_col = low_col)
}

get_PFS_plot <- function(dat, high_col, low_col) {
  surv_dat <- dat |> 
    mutate(time = PFS_time,
           event = PFS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "PFS", high_col = high_col, low_col = low_col)
}

analyse_bac <- function(model_dat,
                        survival_metadata,
                        stool_metabolite_data,
                        serum_metabolite_data,
                        bac,
                        low_col = "grey80",
                        high_col = "grey30") {
  
  joined_bac_dat <- model_dat |> 
    pivot_longer(!pat_id,
                 names_to = "name", 
                 values_to = "value") |> 
    filter(name == bac) |> 
    mutate(name = gsub(".__", "", name),
           name = gsub("_", " ", name)) |> 
    dplyr::inner_join(survival_metadata)
  
  joined_bac_dat <- joined_bac_dat %>%
    mutate(group = if_else(value > 0, "present", "absent")) %>% 
    mutate(group = factor(group, levels = c("absent", "present"))) %>% 
    mutate(cutpoint = 0) %>% 
    mutate(cutpoint =  paste0(cutpoint, "%"))
  
  # Get KM curves
  message("Plot OS")
  km_os_plt <- get_OS_plot(joined_bac_dat, high_col = high_col, low_col = low_col)
  message("Plot PFS")
  km_pfs_plt <- get_PFS_plot(joined_bac_dat, high_col = high_col, low_col = low_col)
  
  # Get Metabolite plots
  message("Plot Stool Metabolites")
  test_metabolites <- c("Valeric acid",
                        "Isovaleric acid",
                        "Indole-3-acetic acid",
                        "Indole-3-carboxyaldehyd")
  stool_metabolite_plots <- test_metabolites |> 
    map(~get_met_plot(bac_dat = joined_bac_dat, 
                      met_dat = stool_metabolite_data,
                      selected_metabolite = .x))
  
  
  message("Plot Serum Metabolites")
  serum_metabolite_plots <- test_metabolites |> 
    map(~get_met_plot(bac_dat = joined_bac_dat, 
                      met_dat = serum_metabolite_data,
                      selected_metabolite = .x,
                      units = "nmol/ml"))
  
  out_plots <- list("KM_OS" = km_os_plt,
                    "KM_PFS" = km_pfs_plt,
                    "VA_stool" = stool_metabolite_plots[[1]],
                    "IVA_stool" = stool_metabolite_plots[[2]],
                    "IAA_stool" = stool_metabolite_plots[[3]],
                    "ICA_stool" = stool_metabolite_plots[[4]],
                    "VA_serum" = serum_metabolite_plots[[1]],
                    "IVA_serum" = serum_metabolite_plots[[2]],
                    "IAA_serum" = serum_metabolite_plots[[3]],
                    "ICA_serum" = serum_metabolite_plots[[4]])
  
  return(out_plots)
}

aggregate_tax_data_acc_to_string <- function(bac_string, 
                                             tax_data) {
  
  message("selecting taxa level")
  
  cat_string <- sub("(_.*)", "", bac_string)
  tax_level <- case_when(cat_string == "s" ~ "Species",
                         cat_string == "g" ~ "Genus",
                         cat_string == "f" ~ "Family",
                         cat_string == "o" ~ "Order",
                         cat_string == "c" ~ "Class",
                         cat_string == "p" ~ "Phylum",
                         cat_string == "k" ~ "Kingdom")
  aggregated_data <- tax_data |> 
    select(pat_id, all_of(tax_level), abundance) |> 
    group_by(!!sym(tax_level), pat_id) %>% 
    summarise(abundance = sum(abundance)) %>% 
    pivot_wider(names_from = !!sym(tax_level), values_from = abundance)
  
  return(aggregated_data)
}
```

#### Plot Significant Described Species
```{r}
sig_species <- results_described_species %>% 
  filter(p.value < 0.05) %>% 
  mutate(Species = ifelse(is.na(Species), Genus, Species)) %>% 
  pull(Species)

plots_species <- sig_species %>% 
  map(function(x) {
    model_dat <- aggregate_tax_data_acc_to_string(x,
                                                  tax_data)
    
    plots <- analyse_bac(model_dat = model_dat,
                         survival_metadata = clinic_data_bac %>% 
                           select(pat_id, PFS_time, PFS_event, OS_time, OS_event) ,
                         stool_metabolite_data = metabo_data_stool,
                         serum_metabolite_data = metabo_data_serum,
                         bac = x,
                         high_col = "red4",
                         low_col = "grey")
    return(plots)
  })

saveRDS(plots_species[[1]]$KM_PFS, file = paste0(fig1dir, "/03_describedSpecies_KM.rds"))
```



### ROC-Curve Described Species & Genera
```{r}
data <- results_described_species %>% 
  dplyr::left_join(pub_data) %>% 
  mutate(score = estimate,
         estimate = ifelse(estimate < 0, "neg", "pos")) %>% 
  ungroup() %>% 
  select(estimate, Direction) %>% 
  dplyr::rename(found_dir = estimate, expected_dir = Direction)

data <- data %>% 
  filter(!is.na(expected_dir))


# Convert to binary for ROC analysis (positive = 1, negative = 0)
# For predictions (found_dir)
data$found_binary <- ifelse(data$found_dir == "pos", 1, 0)
# For actual values (expected_dir)
data$expected_binary <- ifelse(data$expected_dir == "pos", 1, 0)

# Calculate confusion matrix values
TP <- sum(data$found_binary == 1 & data$expected_binary == 1)
FP <- sum(data$found_binary == 1 & data$expected_binary == 0)
TN <- sum(data$found_binary == 0 & data$expected_binary == 0)
FN <- sum(data$found_binary == 0 & data$expected_binary == 1)

# Display confusion matrix
conf_matrix <- matrix(c(TP, FP, FN, TN), nrow = 2,
                      dimnames = list(
                        c("Predicted Positive", "Predicted Negative"),
                        c("Actual Positive", "Actual Negative")
                      ))

# Calculate metrics
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
accuracy <- (TP + TN) / (TP + TN + FP + FN)
roc_obj <- roc(data$expected_binary, data$found_binary)
auc_value <- pROC::auc(roc_obj)

roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

# Create Plot
plt_roc <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "black", size = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  annotate("text", x = 0.05, y = 0.9, label = paste("AUC =", round(auc_value, 3)), hjust = 0, size = 9/.pt) +
  labs(title = "Effect Direction Published vs this Study",
       # subtitle = paste("AUC =", round(auc_value, 3)),
       x = "1 - Specificity",
       y = "Sensitivity"
  ) +
  theme_prism(base_size = 9, border = T) +
  coord_equal()

plt_roc

saveRDS(plt_roc, file = paste0(fig1dir, "/04_coxPH_roc.rds"))
```

### Bray Curtis Heatmap
```{r}
metadata <- meta(physeq)
bray_dist <- phyloseq::distance(physeq, method = "bray")

median(bray_dist)

plt_bray_hm <- pheatmap::pheatmap(as.matrix(bray_dist),
                   show_colnames = FALSE, # No colnames
                   show_rownames = FALSE, # No rownames
                   clustering_distance_rows = as.dist(bray_dist),
                   clustering_distance_cols = as.dist(bray_dist),
                   treeheight_row = 0, # No dendrogram
                   treeheight_col = 0, # No dendrogram
                   main = "Bray Curtis Distance Between Species Abundance in Patients",
                   fontsize = 9
)

plt_bray_hm <- ggplotify::as.ggplot(plt_bray_hm)

saveRDS(plt_bray_hm, file = paste0(fig1dir, "/05_bray_hm.rds"))


```

### Pathways 
```{r}

# Prep
pw_mapping <- pw_data |> 
  group_by(pathway) |> 
  select(pat_id, pathway, cpm) %>% 
  pivot_wider(names_from = pat_id, values_from = cpm) %>% 
  ungroup() %>% 
  mutate(pwID = paste0("pathway",1:n())) %>% 
  select(pathway, pwID)

pw_map <- pw_mapping$pathway
names(pw_map) <- pw_mapping$pwID

model_dat <- pw_data |> 
  mutate(pathway = factor(pathway, labels = names(pw_map))) %>% 
  select(pat_id, pathway, cpm) |>  
  pivot_wider(names_from = pathway, values_from = cpm)

surv_dat <- model_dat |> 
  dplyr::left_join(clinic_data_bac |>
  select(pat_id, PFS_time, PFS_event, OS_time, OS_event) |> 
  mutate(time = PFS_time,
         event = PFS_event))


# Perform Univariate Analysis. Based on https://www.sthda.com/english/wiki/cox-proportional-hazards-model
covariates <- colnames(model_dat)[-1]
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(time, event)~', x)))

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = surv_dat)})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                         x <- summary(x)
                         p.value<-signif(x$wald["pvalue"], digits=2)
                         wald.test<-signif(x$wald["test"], digits=2)
                         beta<-signif(x$coef[1], digits=2);#coeficient beta
                         HR <-signif(x$coef[2], digits=2);#exp(beta)
                         HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                         HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                         HRtotal <- paste0(HR, " (", 
                                      HR.confint.lower, "-", HR.confint.upper, ")")
                         res<-c(beta, HR, HR.confint.lower, HR.confint.upper, HRtotal, wald.test, p.value)
                         names(res)<-c("beta", "HR" , "conf.upper", "conf.lower",
                                       "HR (95% CI for HR)", "wald.test", 
                                       "p.value")
                         return(res)
                         #return(exp(cbind(coef(x),confint(x))))
                       })

uni_cox_res <- t(as.data.frame(univ_results, check.names = FALSE)) |> 
  as_tibble(rownames = "pwID") |> 
  arrange(p.value) 

uni_cox_res$pathway <- pw_map[uni_cox_res$pwID]
  
plt <- uni_cox_res %>% 
    mutate(HR = as.numeric(HR),
           conf.lower = as.numeric(conf.lower),
           conf.upper = as.numeric(conf.upper),
           p.value = as.numeric(p.value),
           dir = ifelse(HR < 1, "down", "up")) %>%
    filter(!is.na(dir)) %>% 
    group_by(dir) %>% 
    slice_min(order_by = p.value, n = 10) %>% 
    ggplot(aes(x = HR, y = reorder(pathway, HR), color = -log10(p.value))) +
    geom_vline(xintercept = 1, color = "black") +
    geom_pointrange(aes(xmin = conf.lower, xmax = conf.upper)) +
    scale_color_gradient(high = "red", low = "black") +
    lims(x = c(0,4)) +
    labs(y = "",
         x = "Hazard Ratio",
         title = "HR according to univariate CoxPH",
         caption = "Top 10 most significant positive and negative associations, 95% CI shown") +
    theme_prism(base_size = 9) +
    theme(legend.title = element_text())
  
  
  
saveRDS(plt, file = paste0(fig1dir, "/06_CoxPH_pw.rds"))
```


## Figure S3 - Previous Species and Metagenome Differences

### Set Paths and Load Data 
```{r}
figS3dir <- "../2_analysis/Figures/FigS3"
dir.create(figS3dir, recursive = T)
```


### OS-KM, Serum and Stool Metabolites in Described Species

```{r}
# From Figure 1 - Analyze Selected Bacteria

saveRDS(plots_species[[1]], file = paste0(figS3dir, "/01ff_Selected_Bacteria_plots.rds"))
```


### Plot Other Species
```{r}

plots_other_species <- c("s__Ruminococcus_gnavus", "s__Akkermansia_muciniphila") %>% 
  map(function(x) {
    model_dat <- aggregate_tax_data_acc_to_string(x,
                                                  tax_data)
    
    plots <- analyse_bac(model_dat = model_dat,
                         survival_metadata = clinic_data_bac %>% 
                           select(pat_id, PFS_time, PFS_event, OS_time, OS_event) ,
                         stool_metabolite_data = metabo_data_stool,
                         serum_metabolite_data = metabo_data_serum,
                         bac = x,
                         high_col = "lightseagreen",
                         low_col = "grey")
    return(plots)
  })

saveRDS(plots_other_species, file = paste0(figS3dir, "/02_Rgnavus_Amuciniphila_plots.rds"))

```


### Bray Curtis NMDS 
```{r}

metadata <- data.frame(sample_data(physeq))
metadata$response_CR_d180 <- as.numeric(metadata$response_CR_d180)

dist_mat <- phyloseq::distance(physeq, method="bray")
set.seed(123)
perm_res <- adonis2(dist_mat ~ response_CR_d180,
       data = metadata, permutations = 999)

# Extract PERMANOVA stats
R2 <- perm_res$R2[1]       # Model R2
Fval <- perm_res$F[1]      # F statistic
pval <- perm_res$`Pr(>F)`[1]  # p-value

# Format the annotation string
annot_text <- sprintf("PERMANOVA:\nR² = %.3f\nF = %.2f\np = %.3f", R2, Fval, pval)

ordination <- ordinate(physeq, method = "NMDS", distance = "bray")
plt_bray_nmds <- plot_ordination(physeq, ordination, color = "response_CR_d180") +
  theme_prism(base_size = 9) +
  scale_color_manual(values = c("hotpink", "green"), labels = c("noCR", "CR")) +
  annotate("text", x = 1.2, y = -1.3, label = annot_text, hjust = 0, size = 9/.pt)
plt_bray_nmds

saveRDS(plt_bray_nmds, file = paste0(figS3dir, "/03_bray_nmds.rds"))

```


### PCoA and PCA Comparison

```{r}
# This requires clustering from Figure 2 - Metabolite Heatmap
# Will be run at a later point in the script
```

## Figure 2 - Microbial Metabolites are associated with Survival
```{r}
fig2dir <- "../2_analysis/Figures/Fig2"
dir.create(fig2dir, recursive = T)
```

### Metabolite Heatmap 

#### Function
```{r}
plot_cluster_heatmap <- function(dat, 
                                 metadata,
                                 clust_met = "ward.D2",
                                 dist_met = "euclidean", 
                                 cluster_order = c("1","2","3","4"),
                                 name_order = c("1","2","3","4")){
  
  mat <- dat %>% select(-pat_id) %>% as.matrix
  rownames(mat) <- dat$pat_id
  
  # Transform Data back and Winsorize data to remove outliers
  hm_dat <- apply(2^mat, MARGIN = 2, DescTools::Winsorize)
  
  # Order for current response
  ord <- order(metadata$response_CR_d180)
  pat_response_ord <- metadata[ord,]
  
  # Extract annotation data
  col_annot_dat <- data.frame(row.names = pat_response_ord$pat_id,
                              response_d180 = as.factor(pat_response_ord$response_CR_d180))
  
  
  # Get treedata from pheatmap
  p <- pheatmap::pheatmap(t(hm_dat[ord,]),
                          scale = "row",
                          cluster_cols = T,
                          show_colnames = F, 
                          clustering_method = clust_met,
                          clustering_distance_cols = dist_met)
  
  # cut for 4 clusters
  met_cluster <- sort(cutree(p$tree_col, k=4))
  
  # add cluster to annotation df 
  clust_df <- data.frame(pat_id = names(met_cluster),
                         met_cluster = met_cluster) %>% 
    mutate(met_cluster = factor(met_cluster, levels = name_order, labels = c("Beneficial1", "Beneficial2", "Beneficial3", "Detrimental")))
  
  
  clust_annot_dat <- col_annot_dat |> 
    as_tibble(rownames = "pat_id") |> 
    left_join(clust_df) |> 
    dplyr::select(-pat_id) |> 
    mutate(response_d180 = as.factor(response_d180),
           response_d180 = fct_recode(response_d180, "CR" = "1", "noCR" = "0")) |> 
    as.data.frame()

  
  rownames(clust_annot_dat) <- rownames(col_annot_dat)
  
  
  annotation_colors <- list(response_d180 = c("noCR" = "hotpink", "CR" = "green2"),
                            met_cluster=c("Beneficial1"="steelblue3",
                                          "Beneficial2"="steelblue1",
                                          "Beneficial3" = "lightskyblue",
                                          "Detrimental"="sienna3"))
  

if (cluster_order[1] == "mirror") {

  callback = function(hc, mat){
    dend = as.dendrogram(hc)

    # Mirror the dendrogram
    dend = rev(dend)

    as.hclust(dend)
  }
} else {
  callback = function(hc, mat){
    # Cut tree into 4 clusters
    k = 4
    clust <- cutree(hc, k = k)

    # Desired cluster display order (left to right)
    desired_order <- cluster_order

    # Get column indices per cluster in desired order
    ord <- unlist(lapply(desired_order, function(cl) {
      which(clust == cl)
    }))

    # Reorder dendrogram manually
    dend <- as.dendrogram(hc)
    dend <- reorder(dend, wts = ord)

    # Return reordered hclust
    as.hclust(dend)
  }
}

  # Plot Heatmap
  plt_hm  <- pheatmap::pheatmap(t(hm_dat[ord,]),
                             scale = "row",
                             cluster_cols = T,
                             annotation_col = clust_annot_dat,
                             annotation_colors = annotation_colors,
                             show_colnames = F,
                             cutree_rows = 4, 
                             cutree_cols = 4,
                             height = 10,
                             width = 14,
                             border_color = NA,
                             # filename = paste0(outdir,"/cluster_heatmap.png"),
                             clustering_method = clust_met,
                             clustering_distance_cols = dist_met, 
                             clustering_callback = callback,
                             fontsize = 9,
                             treeheight_row = 0, # No dendrogram
                             treeheight_col = 0, # No dendrogram
                             annotation_legend = F
  )
  
  plt_hm <- ggplotify::as.ggplot(plt_hm)
  
  return(list("cluster" = clust_df,
              "plot" = plt_hm))
}
```

#### Plot
```{r}
hm_dat_stool <- plot_cluster_heatmap(metabo_data_stool,
                            clinic_data_stool, 
                            name_order = c("4", "3", "2", "1"),
                            cluster_order = c( "3", "2", "1", "4"))


plt <- hm_dat_stool$plot


saveRDS(plt, file = paste0(fig2dir, "/01_HM_stool.rds"))
```


### Metabolite Cluster Survival Curves

```{r}
plot_metcluster_survcurves <- function(metadata, 
                                       cluster_data, 
                                       event_type) {
  surv_dat2 <- metadata |> 
    left_join(cluster_data) %>%  
    mutate(clust_cat = ifelse(met_cluster == "Detrimental", "Detrimental", "Beneficial")) %>% 
    mutate(clust_cat = factor(clust_cat, levels = c("Beneficial", "Detrimental")))
  
  if(event_type == "PFS") {
    surv_dat2 <- surv_dat2 |> 
      mutate(event = PFS_event,
             time = PFS_time)
  } else if(event_type == "OS") {
    surv_dat2 <- surv_dat2 |> 
      mutate(event = OS_event,
             time = OS_time)
  }
  
  surv_fit <- survminer::surv_fit(Surv(time, event) ~ clust_cat, data = surv_dat2)
  LR_test <- survdiff(Surv(time, event) ~ clust_cat, data = surv_dat2)
  cox_fit <- coxph(Surv(time, event) ~ clust_cat, data = surv_dat2)
  sumcox <- summary(cox_fit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  time_cutoff = max(surv_dat2$time)
  survs = summary(surv_fit, times=time_cutoff)$surv
  
  surv <- ggsurvplot(surv_fit,
                     pval = FALSE,
                     conf.int = T,
                     risk.table = TRUE,
                     palette = c( "steelblue3", "sienna3" ),
                     fontsize = 9/.pt)
  
  p1 <- surv$plot +
    labs(y = paste0("Probablity of ", event_type, " (%)"),
         x = "Days after CAR-T") +
    scale_color_manual(values = c( "steelblue3", "sienna3" ), label = c( "Beneficial", "Detrimental"))+
    scale_fill_manual(values = c(  "steelblue3", "sienna3" ), label = c("Beneficial", "Detrimental")) +
    theme_prism(base_size = 9) +
    labs(subtitle = paste0(
      # "cox PH: p = ",
      # signif(sumcox$waldtest[3], 2),
      "p = ",
      signif(LR_test$pvalue, 2),
      ", HR = ",
      signif(exp(cox_fit$coefficients), 3),
      " (",HR.confint.lower,"-",HR.confint.upper,")")) +
    theme(legend.position = "top")  +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  plot(p1)
  
  tp <- surv$table + 
    theme_prism(base_size = 9) +
    scale_color_manual(values = c( "steelblue3",  "sienna3" ), label = c("Beneficial", "Detrimental"))+
    scale_fill_manual(values = c(  "steelblue3", "sienna3" ), label = c("Beneficial", "Detrimental")) +
    scale_y_discrete(labels = c("Detrimental", "Beneficial")) +
    theme(axis.text.y = element_text(color =c("sienna3", "steelblue3"))) +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
    rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    theme(plot.title = element_text(hjust = 0, size=9)) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  p2 <- ggarrange(p1, tp, ncol = 1, heights = c(3,1))  
  p2 <- (free(p1, type = "label") / tp) + plot_layout(heights = c(5, 1))
  return(p2)
}

plt_clust_PFS<- plot_metcluster_survcurves(clinic_data_stool, hm_dat_stool$cluster, "PFS")
plot(plt_clust_PFS)

plt_clust_OS<- plot_metcluster_survcurves(clinic_data_stool, hm_dat_stool$cluster, "OS")
plot(plt_clust_OS)


saveRDS(plt_clust_PFS, file = paste0(fig2dir, "/02_KM_Metcluster_stool_PFS.rds"))
saveRDS(plt_clust_OS, file = paste0(fig2dir, "/03_KM_Metcluster_stool_OS.rds"))
```

### Metabolite Survival

#### Function
```{r}
create_KM_plot <- function(survival_dat,
                           event_type,
                           show_n = TRUE,
                           risk_table = FALSE,
                           low_col = "grey80",
                           high_col = "grey30",
                           breaks.by = 200) {
  
  # Create survival fit
  fit <- survminer::surv_fit(Surv(time, event) ~ group, data = survival_dat)
  LR_test <- survdiff(Surv(time, event) ~ group, data = survival_dat)
  coxfit <- coxph(Surv(time, event) ~ group, data = survival_dat)
  sumcox <- summary(coxfit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  caption <- paste0("cutpoint:  <= ", survival_dat$cutpoint)
  
  time_cutoff = max(survival_dat$time)
  survs = summary(fit, times=time_cutoff)$surv
  if(event_type == "PFS") {
    time.breaks <- diff((scales::extended_breaks())(fit$time))[1]
  } else if(event_type == "OS") {
    time.breaks <- breaks.by
  }
  
  # Generate survival plot
  surv <- ggsurvplot(fit,
                     pval = FALSE,          # Show p-value
                     conf.int = FALSE,     # Don't show confidence intervals
                     risk.table = TRUE,    # Show risk table
                     size = 1.5,
                     fontsize = 9/.pt,
                     break.time.by = time.breaks)
  
  # Customize KM plot appearance
  p1 <- surv$plot 
  
  if (show_n == TRUE) {
    p1 <- p1 + scale_color_manual(values = c(low_col, high_col),
                                  labels = c(paste0(levels(survival_dat$group)[1], " n = ", table(survival_dat$group)[1]), 
                                             paste0(levels(survival_dat$group)[2], " n = ", table(survival_dat$group)[2])))
  } else if (show_n == FALSE) {
    p1 <- p1 + scale_color_manual(values = c(low_col, high_col),
                                  labels = c(levels(survival_dat$group)[1], 
                                             levels(survival_dat$group)[2]))
  }
  
  p1 <- p1 +
    labs(y = paste0("Probability of ", event_type , " (%)"),
         x = "Days after CAR-T",
         title = survival_dat$name[1],
         subtitle = paste0(
           # "cox PH: p = ",
           # signif(sumcox$waldtest[3], 2),
           "p = ",
           signif(LR_test$pvalue, 2),
           ", HR = ",
           signif(exp(coxfit$coefficients), 2),
           " (",HR.confint.lower,"-",HR.confint.upper,")"),
         caption = caption
    ) +
    theme_prism(base_size = 9) +
    guides(colour = guide_legend(position = "inside")) +
    theme(legend.position.inside = c(0.8, 0.12),
          legend.key.height = unit(10, "pt")) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  if (risk_table == TRUE) {
    p2 <- surv$table + 
      theme_prism(base_size = 9) +
      scale_color_manual(values = c(low_col, high_col), label = c(levels(survival_dat$group)[1], levels(survival_dat$group)[2]))+
      scale_fill_manual(values = c(low_col, high_col), label = c(levels(survival_dat$group)[1], levels(survival_dat$group)[2])) +
      scale_y_discrete(labels = c(levels(survival_dat$group)[2], levels(survival_dat$group)[1])) +
      theme(axis.text.y = element_text(color =c(high_col,  low_col))) +
          rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
    
    p1 <- (free(p1, type = "label") / p2) + plot_layout(heights = c(5, 1))
      
  }
  
  
  return(list("plot" = p1,
              "breaks" = time.breaks))
}

get_OS_plot <- function(dat, ...) {
  surv_dat <- dat |> 
    mutate(time = OS_time,
           event = OS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "OS", ...)
}

get_PFS_plot <- function(dat, ...) {
  surv_dat <- dat |> 
    mutate(time = PFS_time,
           event = PFS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "PFS", ...)
}


get_cutoff_groups <- function(data,
                              grouping_type = "surv_cutoff",
                              custom_cutoff = NULL) {
  
  if( grouping_type == "custom") {
    grouped_dat <- data %>%
      mutate(group = if_else(value > custom_cutoff, "high", "low")) %>% 
      mutate(group = factor(group, levels = c("low", "high"))) %>% 
      mutate(cutpoint = custom_cutoff)
  } else if ( grouping_type == "surv_cutoff") {
    cutpoint_fit <- surv_cutpoint(data,
                                  time = "PFS_time",
                                  event = "PFS_event",
                                  minprop = 0.1,  # Minimum proportion in each group
                                  variables = "value")
    
    cutpoint <- summary(cutpoint_fit)[1, 1]
    grouped_dat <- data %>%
      mutate(group = if_else(value > cutpoint, "high", "low")) %>% 
      mutate(group = factor(group, levels = c("low", "high"))) %>% 
      mutate(cutpoint = cutpoint)
  }
  
  return(grouped_dat)
}


analyse_metabolite <- function(dat,
                               metadata,
                               metabolite_name,
                               grouping_type = "surv_cutoff",
                               custom_cutoff = 0,
                               unit = "µmol/g",
                               high_col,
                               low_col, 
                               type = "both") {
  surv_dat <- dat %>% 
    pivot_longer(!pat_id,
                 names_to = "name",
                 values_to = "value") %>% 
    filter(name == metabolite_name) %>% 
    left_join(metadata %>% select(pat_id,
                                  PFS_time, PFS_event,
                                  OS_time, OS_event, 
                                  response_CR_d180)) %>%
    mutate(value = 2^value)
  
  grouped_surv_dat <- get_cutoff_groups(surv_dat,
                                        grouping_type = grouping_type,
                                        custom_cutoff = custom_cutoff) %>% 
    # mutate(cutpoint = 2^cutpoint) %>% 
    mutate(cutpoint_orig = cutpoint,
           cutpoint = paste0(signif(cutpoint, digits = 2), " ", unit))
  
  
  
  km_pfs_plt <- get_PFS_plot(grouped_surv_dat, high_col = high_col, low_col = low_col)
  pfs_breaks <- km_pfs_plt$breaks
  km_pfs_plt <- km_pfs_plt$plot
  km_os_plt <- get_OS_plot(grouped_surv_dat, high_col = high_col, low_col = low_col, breaks.by = pfs_breaks)
  
  if (type == "main_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE)
    return(km_pfs_plt$plot)
  } else if (type == "sup_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE,
                               risk_table = TRUE)
    km_os_plt <- get_OS_plot(grouped_surv_dat,
                             high_col = high_col,
                             low_col = low_col,
                             show_n = FALSE,
                             risk_table = TRUE,
                             breaks.by = pfs_breaks)
    return(ggarrange(km_pfs_plt$plot,km_os_plt$plot, nrow = 1))
  } else if (type == "cutpoint") {
    return(grouped_surv_dat$cutpoint_orig[1])
  } else if (type == "serum_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE,
                               risk_table = TRUE)
    return(km_pfs_plt$plot)
  }
  
}
```

#### SCFA Plot 
```{r}
metabolites <- c("Acetic acid",
                "Propionic acid",
                "Butyric acid",
                "Valeric acid")

plts_SCFA <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "main_figure")
  })

saveRDS(plts_SCFA, file = paste0(fig2dir, "/04_SCFA_stool_main.rds"))
```


#### BCFA Plot
```{r}
metabolites <- c("Isobutyric acid",
                "Isovaleric acid")

plts_BCFA <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "main_figure")
  })

saveRDS(plts_BCFA, file = paste0(fig2dir, "/05_BCFA_stool_main.rds"))

```

#### Indoles Plot

```{r}
metabolites <- c("Indole-3-acetic acid",
                 "Indole-3-carboxyaldehyd")

plts_Indoles <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "main_figure")
  })

saveRDS(plts_Indoles, file = paste0(fig2dir, "/06_Indoles_stool_main.rds"))
```

## Figure S3 - PCoA and PCA

### PCA
```{r}
# requires Figure2 - Metabolite Heatmap

pca_dat <- metabo_data_stool %>% 
  column_to_rownames("pat_id")

pca_dat <-  scale(apply(2^pca_dat, MARGIN = 2, DescTools::Winsorize)) %>%
  as_tibble(rownames = "pat_id")
  
pca_fit <- pca_dat %>% 
  select(where(is.numeric)) %>%
  select(where(~ var(.x, na.rm = TRUE) != 0)) %>%
  prcomp(scale. = F)

expl_var <- summary(pca_fit)$importance[2,]*100


metadata <- pca_fit %>%
  augment(pca_dat %>% 
            select(pat_id) %>% 
            dplyr::left_join(clinic_data_stool) %>%
            dplyr::left_join(hm_dat_stool$cluster)) %>% 
  column_to_rownames("pat_id")

metadata <- metadata %>% mutate(met_cluster = ifelse(met_cluster == "Detrimental","Detrimental","Beneficial"))

dist_mat <- dist(pca_dat %>% column_to_rownames("pat_id"))

set.seed(1234)
perm_res <- adonis2(dist_mat ~ met_cluster, data = metadata, permutations = 1000)

# Extract PERMANOVA stats
R2 <- perm_res$R2[1]       # Model R2
Fval <- perm_res$F[1]      # F statistic
pval <- perm_res$`Pr(>F)`[1]  # p-value

# Format the annotation string
annot_text <- sprintf("Detrimental vs All Benefitial\nPERMANOVA:\nR² = %.3f\nF = %.2f\np = %.3f", R2, Fval, pval)


plt_metabo <- pca_fit %>%
  augment(pca_dat %>% 
            select(pat_id) %>% 
            dplyr::left_join(clinic_data_stool) %>%
            dplyr::left_join(hm_dat_stool$cluster)) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2)) + 
  ggforce::geom_mark_ellipse(aes(color = met_cluster), alpha = 0.3) +
  geom_point(size = 1,
             aes(
               color = as.factor(met_cluster),
                 shape = as.factor(response_CR_d180)),
             alpha = 0.8) +
  scale_color_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
    scale_fill_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
  scale_shape_manual(values = c(3,16), labels = c("noCR", "CR")) +
  labs(x = paste0("PC1 [", signif(expl_var[1], 2),"%]"),
       y = paste0("PC2 [", signif(expl_var[2], 2),"%]")) +
  annotate("text", x = -7, y = -4.5, label = annot_text, hjust = 0, vjust = 1, size = 9/.pt) +
  theme_prism(base_size = 9)

saveRDS(plt_metabo, file = paste0(figS3dir, "/04_PCA_Metabolomics.rds"))

```


### PCoA
```{r}
# Generate physeq object with cluster information
sample_df <- clinic_data_bac |>
  dplyr::left_join(hm_dat_stool$cluster) %>% 
  dplyr::select(response_CR_d180,
                met_cluster) |> 
  mutate(response_CR_d180 = as.factor(response_CR_d180)) |> 
  as.data.frame()

rownames(sample_df) <- clinic_data_bac$pat_id

SAMPLE <- sample_data(object = sample_df)

physeq_PCoA <- phyloseq(OTU,TAX,SAMPLE)
saveRDS(physeq_PCoA, file = "../2_analysis/02_analysisData/physeq_PCoA.rds")

metadata <- data.frame(sample_data(physeq_PCoA))
metadata <- metadata %>% mutate(met_cluster = ifelse(is.na(met_cluster),"NA",met_cluster))
metadata <- metadata %>% mutate(met_cluster = ifelse(met_cluster == "4","Detrimental","Beneficial"))
dist_mat <- phyloseq::distance(physeq_PCoA, method="bray")
set.seed(123)
perm_res <- adonis2(dist_mat ~ met_cluster,
       data = metadata, permutations = 1000)
perm_res

# Extract PERMANOVA stats
R2 <- perm_res$R2[1]       # Model R2
Fval <- perm_res$F[1]      # F statistic
pval <- perm_res$`Pr(>F)`[1]  # p-value

# Format the annotation string
annot_text <- sprintf("Detrimental vs All Benefitial\nPERMANOVA:\nR² = %.3f\nF = %.2f\np = %.3f", R2, Fval, pval)


# Plot PCoA

ordination <- ordinate(physeq_PCoA, method = "PCoA", distance = "bray")



plt_metage <- plot_ordination(physeq_PCoA, ordination, justDF = T) %>% 
  ggplot(aes(x = Axis.1, y = Axis.2)) +
  ggforce::geom_mark_ellipse(aes(color = met_cluster, shape = NULL), alpha = 0.3) +
  geom_point(size = 1,
             aes(
               color = as.factor(met_cluster),
                 # fill = as.factor(met_cluster),
                 shape = as.factor(response_CR_d180)),
             alpha = 0.8) +
  scale_color_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
    scale_fill_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
  scale_shape_manual(values = c(3,16), labels = c("noCR", "CR")) +
  labs(x = paste0("PCo1 [",signif(ordination$values$Relative_eig[1],3)*100,"%]"),
       y = paste0("PCo2 [",signif(ordination$values$Relative_eig[2],3)*100,"%]")) +
  annotate("text", x = -0.5, y = -0.25, label = annot_text, hjust = 0, vjust = 1, size = 9/.pt) +
    theme_prism(base_size = 9)

saveRDS(plt_metage, file = paste0(figS3dir, "/05_PCoA_Metagenomics.rds"))

# NMDS 

ordination <- ordinate(physeq_PCoA, method = "NMDS", distance = "bray")

plt_metage_nmds <- plot_ordination(physeq_PCoA, ordination, justDF = T) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  ggforce::geom_mark_ellipse(aes(color = met_cluster, shape = NULL), alpha = 0.3) +
  geom_point(size = 1,
             aes(
               color = as.factor(met_cluster),
                 # fill = as.factor(met_cluster),
                 shape = as.factor(response_CR_d180)),
             alpha = 0.8) +
  scale_color_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
    scale_fill_manual(
    values = c( "steelblue3", "steelblue1", "lightskyblue" , "sienna3" ),
  )  +
  scale_shape_manual(values = c(3,16), labels = c("noCR", "CR")) +
  labs(x = "NMDS1",
       y = "NMDS2") +
    theme_prism(base_size = 9)

plt_metage_nmds
```


## Figure S4 - Influence of Metabolites in Stool on Survival

### Set Paths and Load Data 
```{r}
figS4dir <- "../2_analysis/Figures/FigS4"
dir.create(figS4dir, recursive = T)
```

### SCFA
```{r}
# Requires functions from Figure 2 - Metabolite Survival

metabolites <- c("Acetic acid",
                "Propionic acid",
                "Butyric acid",
                "Valeric acid")

plts_SCFA <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "sup_figure")
  })

saveRDS(plts_SCFA, file = paste0(figS4dir, "/01_KM_SCFA_stool_supplement.rds"))
```

### BCFA
```{r}
metabolites <- c("Isobutyric acid",
                 "2-Methyl butyric acid",
                 "Isovaleric acid")

plts_BCFA <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "sup_figure")
  })

saveRDS(plts_BCFA, file = paste0(figS4dir, "/02_KM_BCFA_stool_supplement.rds"))
```

### Indoles
```{r}
metabolites <- c("Indole-3-acetic acid",
                 "Indole-3-carboxyaldehyd",
                 "Indole-3-propionic acid")

plts_Indoles <- metabolites %>% 
  map(function(x){
    analyse_metabolite(metabo_data_stool,
                   metadata = clinic_data_stool,
                   metabolite_name = x,
                   unit = "µmol/g",
                   high_col = "black",
                   low_col = "grey", 
                   type = "sup_figure")
  })

saveRDS(plts_Indoles, file = paste0(figS4dir, "/03_KM_Indoles_stool_supplement.rds"))
```

## Figure 3 - Clustering Serum Metabolites 

#### Function
```{r}
create_KM_plot <- function(survival_dat,
                           event_type,
                           show_n = TRUE,
                           risk_table = FALSE,
                           low_col = "grey80",
                           high_col = "grey30",
                           breaks.by = 200) {
  
  # Create survival fit
  fit <- survminer::surv_fit(Surv(time, event) ~ group, data = survival_dat)
  LR_test <- survdiff(Surv(time, event) ~ group, data = survival_dat)
  coxfit <- coxph(Surv(time, event) ~ group, data = survival_dat)
  sumcox <- summary(coxfit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  caption <- paste0("cutpoint:  <= ", survival_dat$cutpoint)
  
  time_cutoff = max(survival_dat$time)
  survs = summary(fit, times=time_cutoff)$surv
  if(event_type == "PFS") {
    time.breaks <- diff((scales::extended_breaks())(fit$time))[1]
  } else if(event_type == "OS") {
    time.breaks <- breaks.by
  }
  
  # Generate survival plot
  surv <- ggsurvplot(fit,
                     pval = FALSE,          # Show p-value
                     conf.int = FALSE,     # Don't show confidence intervals
                     risk.table = TRUE,    # Show risk table
                     size = 1.5,
                     fontsize = 9/.pt,
                     break.time.by = time.breaks)
  
  # Customize KM plot appearance
  p1 <- surv$plot 
  
  if (show_n == TRUE) {
    p1 <- p1 + scale_color_manual(values = c(low_col, high_col),
                                  labels = c(paste0(levels(survival_dat$group)[1], " n = ", table(survival_dat$group)[1]), 
                                             paste0(levels(survival_dat$group)[2], " n = ", table(survival_dat$group)[2])))
  } else if (show_n == FALSE) {
    p1 <- p1 + scale_color_manual(values = c(low_col, high_col),
                                  labels = c(levels(survival_dat$group)[1], 
                                             levels(survival_dat$group)[2]))
  }
  
  p1 <- p1 +
    labs(y = paste0("Probability of ", event_type , " (%)"),
         x = "Days after CAR-T",
         title = survival_dat$name[1],
         subtitle = paste0(
           # "cox PH: p = ",
           # signif(sumcox$waldtest[3], 2),
           "p = ",
           signif(LR_test$pvalue, 2),
           ", HR = ",
           signif(exp(coxfit$coefficients), 2),
           " (",HR.confint.lower,"-",HR.confint.upper,")"),
         caption = caption
    ) +
    theme_prism(base_size = 9) +
    guides(colour = guide_legend(position = "inside")) +
    theme(legend.position.inside = c(0.8, 0.12),
          legend.key.height = unit(10, "pt")) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  if (risk_table == TRUE) {
    p2 <- surv$table + 
      theme_prism(base_size = 9) +
      scale_color_manual(values = c(low_col, high_col), label = c(levels(survival_dat$group)[1], levels(survival_dat$group)[2]))+
      scale_fill_manual(values = c(low_col, high_col), label = c(levels(survival_dat$group)[1], levels(survival_dat$group)[2])) +
      scale_y_discrete(labels = c(levels(survival_dat$group)[2], levels(survival_dat$group)[1])) +
      theme(axis.text.y = element_text(color =c(high_col,  low_col))) +
          rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
    
    p1 <- (free(p1, type = "label") / p2) + plot_layout(heights = c(5, 1))
      
  }
  
  
  return(list("plot" = p1,
              "breaks" = time.breaks))
}

get_OS_plot <- function(dat, ...) {
  surv_dat <- dat |> 
    mutate(time = OS_time,
           event = OS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "OS", ...)
}

get_PFS_plot <- function(dat, ...) {
  surv_dat <- dat |> 
    mutate(time = PFS_time,
           event = PFS_event) |> 
    select(pat_id, name, group, time, event, cutpoint)
  
  create_KM_plot(surv_dat, "PFS", ...)
}


get_cutoff_groups <- function(data,
                              grouping_type = "surv_cutoff",
                              custom_cutoff = NULL) {
  
  if( grouping_type == "custom") {
    grouped_dat <- data %>%
      mutate(group = if_else(value > custom_cutoff, "high", "low")) %>% 
      mutate(group = factor(group, levels = c("low", "high"))) %>% 
      mutate(cutpoint = custom_cutoff)
  } else if ( grouping_type == "surv_cutoff") {
    cutpoint_fit <- surv_cutpoint(data,
                                  time = "PFS_time",
                                  event = "PFS_event",
                                  minprop = 0.1,  # Minimum proportion in each group
                                  variables = "value")
    
    cutpoint <- summary(cutpoint_fit)[1, 1]
    grouped_dat <- data %>%
      mutate(group = if_else(value > cutpoint, "high", "low")) %>% 
      mutate(group = factor(group, levels = c("low", "high"))) %>% 
      mutate(cutpoint = cutpoint)
  }
  
  return(grouped_dat)
}


analyse_metabolite <- function(dat,
                               metadata,
                               metabolite_name,
                               grouping_type = "surv_cutoff",
                               custom_cutoff = 0,
                               unit = "µmol/g",
                               high_col,
                               low_col, 
                               type = "both") {
  surv_dat <- dat %>% 
    pivot_longer(!pat_id,
                 names_to = "name",
                 values_to = "value") %>% 
    filter(name == metabolite_name) %>% 
    left_join(metadata %>% select(pat_id,
                                  PFS_time, PFS_event,
                                  OS_time, OS_event, 
                                  response_CR_d180)) %>%
    mutate(value = 2^value)
  
  grouped_surv_dat <- get_cutoff_groups(surv_dat,
                                        grouping_type = grouping_type,
                                        custom_cutoff = custom_cutoff) %>% 
    # mutate(cutpoint = 2^cutpoint) %>% 
    mutate(cutpoint_orig = cutpoint,
           cutpoint = paste0(signif(cutpoint, digits = 2), " ", unit))
  
  
  
  km_pfs_plt <- get_PFS_plot(grouped_surv_dat, high_col = high_col, low_col = low_col)
  pfs_breaks <- km_pfs_plt$breaks
  km_pfs_plt <- km_pfs_plt$plot
  km_os_plt <- get_OS_plot(grouped_surv_dat, high_col = high_col, low_col = low_col, breaks.by = pfs_breaks)
  
  if (type == "main_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE)
    return(km_pfs_plt$plot)
  } else if (type == "sup_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE,
                               risk_table = TRUE)
    km_os_plt <- get_OS_plot(grouped_surv_dat,
                             high_col = high_col,
                             low_col = low_col,
                             show_n = FALSE,
                             risk_table = TRUE,
                             breaks.by = pfs_breaks)
    return(ggarrange(km_pfs_plt$plot,km_os_plt$plot, nrow = 1))
  } else if (type == "cutpoint") {
    return(grouped_surv_dat$cutpoint_orig[1])
  } else if (type == "serum_figure") {
    km_pfs_plt <- get_PFS_plot(grouped_surv_dat,
                               high_col = high_col,
                               low_col = low_col,
                               show_n = FALSE,
                               risk_table = TRUE)
    return(km_pfs_plt$plot)
  }
  
}
```

### Set Paths and Load Data 
```{r}
fig3dir <- "../2_analysis/Figures/Fig3"
dir.create(fig3dir, recursive = T)
```

### Metabolite Heatmap
```{r}
# Requires function from Figure2 - Metabolite Heatmap

hm_dat_serum <- plot_cluster_heatmap(metabo_data_serum,
                               clinic_data_serum, 
                               name_order = c("4", "3", "2", "1"),
                               cluster_order = "mirror")

plt <- hm_dat_serum$plot


saveRDS(plt, file = paste0(fig3dir, "/01_HM_serum.rds"))

```
### Metabolite Cluster Survival Curves

```{r}
# Requires Function from Figure2 - Metabolite Cluster Survival Curves


plt_clust_PFS <- plot_metcluster_survcurves(clinic_data_serum, hm_dat_serum$cluster, "PFS")
plot(plt_clust_PFS)

plt_clust_OS <- plot_metcluster_survcurves(clinic_data_serum, hm_dat_serum$cluster, "OS")
plot(plt_clust_OS)


saveRDS(plt_clust_PFS, file = paste0(fig3dir, "/02_KM_Metcluster_serum_PFS.rds"))
saveRDS(plt_clust_OS, file = paste0(fig3dir, "/03_KM_Metcluster_serum_OS.rds"))

```


### Individual Metabolites
```{r}

p1 <- analyse_metabolite(metabo_data_serum,
                   metadata = clinic_data_serum,
                   metabolite_name = "Butyric acid",
                   unit = "nmol/ml",
                   high_col = "#CFB1E7",
                   low_col = "grey", 
                   type = "serum_figure")
p2 <- analyse_metabolite(metabo_data_serum,
                   metadata = clinic_data_serum,
                   metabolite_name = "Valeric acid",
                   unit = "nmol/ml",
                   high_col = "#BE97DE",
                   low_col = "grey", 
                   type = "serum_figure")
p3 <- analyse_metabolite(metabo_data_serum,
                   metadata = clinic_data_serum,
                   metabolite_name = "Isovaleric acid",
                   unit = "nmol/ml",
                   high_col = "#CCE4FD",
                   low_col = "grey", 
                   type = "serum_figure")
p4 <- analyse_metabolite(metabo_data_serum,
                   metadata = clinic_data_serum,
                   metabolite_name = "Indole-3-acetic acid",
                   unit = "nmol/ml",
                   high_col = "#FCC384",
                   low_col = "grey", 
                   type = "serum_figure")
plts <- list(p1,p2,p3,p4)

saveRDS(plts, file = paste0(fig3dir, "/04_Metabolites_Serum.rds"))
```



## Figure 6 - Risk Score
### Set Paths and Load Data 
```{r}
fig6dir <- "../2_analysis/Figures/Fig6"
dir.create(fig6dir, recursive = T)
```

### Functions
```{r}
get_cutoff_groups <- function(data) {
  
  # Every group is a factor with the "low" group as the first level
  # surv_cutoff uses PFS to find cutpoints
  cutpoint_fit <- surv_cutpoint(data,
                                  time = "PFS_time",
                                  event = "PFS_event",
                                  minprop = 0.1,  # Minimum proportion in each group
                                  variables = "value")
    
    cutpoint <- summary(cutpoint_fit)[1, 1]
    grouped_dat <- data %>%
      mutate(group = if_else(value > cutpoint, "high", "low")) %>% 
      mutate(group = factor(group, levels = c("low", "high"))) %>% 
      mutate(cutpoint = cutpoint)
  return(grouped_dat)
}

get_metabolite_group <- function(metabolite, data){
  
  met_dat <- data %>% 
    pivot_longer(!pat_id) %>% 
    filter(name == metabolite) %>% 
    mutate(value = 2^value) %>% 
    dplyr::left_join(clinic_data_stool, by = c("pat_id" = "pat_id")) %>% 
    dplyr::select(pat_id, response_CR_d180, PFS_event, PFS_time, value)
  
  group_df <- get_cutoff_groups(met_dat) %>% 
    select(pat_id, group, cutpoint) %>% 
    dplyr::rename(value = group) %>% 
    mutate(name = metabolite)
  
}


plt_risk_score <- function(score_df, metadata, surv_type, binary_score = FALSE) {
  
  surv_dat <- score_df %>% 
    left_join(metadata)
  
  if (surv_type == "PFS") {
    surv_dat <- surv_dat %>% 
      mutate(event = PFS_event, time = PFS_time)  %>% 
      select(pat_id, score, event, time, response_CR_d180)
  } else if (surv_type == "OS") {
    surv_dat <- surv_dat %>% 
      mutate(event = OS_event, time = OS_time)  %>% 
      select(pat_id, score, event, time, response_CR_d180)
  }
  
  pal <- RColorBrewer::brewer.pal(n = 4, "Spectral")
  
  if (binary_score == TRUE) {
    surv_dat <- surv_dat %>% 
      mutate(score = ifelse(score == max(score), "high Risk", "low Risk"),
             score = factor(score, levels = c("low Risk", "high Risk")))

    
    custom_labs <- c("low Risk", "high Risk")
    custom_col <- pal[c(4,1)]
  } else if (binary_score == FALSE) {
    scores <- sort(unique(surv_dat$score))
    custom_labs <- paste0("Risk Score ", scores)
    # custom_labs <- rev(c("Risk Score 4", "Risk Score 3", "Risk Score 2", "Risk Score 1"))
    custom_col <- rev(pal)
  }
  
  coxfit <- coxph(Surv(time, event) ~ score , data = surv_dat)
  LR_test <- survdiff(Surv(time, event) ~ score, data = surv_dat)
  fit <- survminer::surv_fit(Surv(time, event) ~ score, data = surv_dat)
  sumcox <- summary(coxfit)
  
  HR.confint.lower <- signif(sumcox$conf.int[,"lower .95"], 2)
  HR.confint.upper <- signif(sumcox$conf.int[,"upper .95"], 2)
  
  time_cutoff = max(surv_dat$time)
  survs = summary(fit, times=time_cutoff)$surv
  
  surv_plt <- ggsurvplot(fit,
                         pval = FALSE,          # Show p-value
                         conf.int = FALSE,     # Don't show confidence intervals
                         risk.table = TRUE,    # Show risk table
                         size = 1,
                         fontsize = 7/.pt)
  
  
  p1 <- surv_plt$plot +
    scale_color_manual(values = custom_col, labels = custom_labs) +
    labs(y = paste0("Probability of ", surv_type," (%)"),
         x = "Days after CAR-T",
         title = NULL,
         subtitle = paste0("p = ",
                           signif(LR_test$pvalue, 2),
                           ", HR = ",
                           signif(exp(coxfit$coefficients), 2),
                           " (",HR.confint.lower,"-",HR.confint.upper,")"),
    ) +
    theme_prism(base_size = 7, base_fontface = "plain") +
    guides(colour = guide_legend(position = "inside")) +
    theme(legend.position.inside = c(0.8, 0.12),
          legend.key.height = unit(10, "pt")) +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  p2 <- surv_plt$table + 
    theme_prism(base_size = 7) +
    scale_y_discrete(labels = rev(custom_labs)) +
    theme(axis.text.y = element_text(color = rev(custom_col))) +
    labs(caption = NULL,title = NULL, x = NULL, y = NULL) +
        rremove("x.text") +
    rremove("x.ticks") +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks") +
    coord_cartesian(xlim = c(0, time_cutoff), 
                    clip = 'on')
  
  p3 <- ggarrange(p1, p2, ncol = 1, heights = c(3,1))  
  
  p3 <- (free(p1, type = "label") / p2) + plot_layout(heights = c(3, 1))
  
  return(p3)
}


format_cox_table <- function(model, surv_dat_clean, noDigits = 2, refLabel = "reference") {
  
  # Required packages
  requireNamespace("broom", quietly = TRUE)
  
  # Extract data
  data <- eval(model$call$data)
  terms <- attr(model$terms, "dataClasses")[-1]
  
  #browser()
  # Model summaries
  coef <- as.data.frame(broom::tidy(model, conf.int = TRUE))
  gmodel <- broom::glance(model)
  
  # Build term structure
  allTerms <- lapply(seq_along(terms), function(i) {
    var <- names(terms)[i]
    
    if (terms[i] %in% c("factor", "character")) {
      adf <- as.data.frame(table(data[, var]))
      cbind(var = var, adf, pos = seq_len(nrow(adf)))
      
    } else if (terms[i] == "numeric") {
      data.frame(var = var, level = "", N = nrow(data), pos = 1)
      
    } else {
      vars <- grep(paste0("^", var, "*."), coef$term, value = TRUE)
      data.frame(var = vars, level = "", N = nrow(data), pos = seq_along(vars))
    }
  })
  
  # Standardize column names
  for (i in seq_along(allTerms)) {
    colnames(allTerms[[i]]) <- c("var", "level", "N", "pos")
  }
  
  allTermsDF <- do.call(rbind, allTerms)
  
  # Match coefficients
  inds <- apply(allTermsDF[, c("var", "level")], 1, paste0, collapse = "")
  coef$term <- gsub("`", "", coef$term)
  rownames(coef) <- coef$term
  
  toShow <- cbind(allTermsDF, coef[inds, ])[, 
    c("var", "level", "N", "p.value", "estimate", "conf.low", "conf.high", "pos")
  ]
  
  # Exponentiate estimates
  toShowExp <- exp(toShow[, c("estimate", "conf.low", "conf.high")])
  toShowExp <- format(toShowExp, digits = noDigits)
  
  # Combine
  out <- data.frame(
    toShow,
    pvalue = signif(toShow$p.value, noDigits + 1),
    toShowExp
  )
  
  # Formatting
  out$stars <- paste0(
    round(out$p.value, noDigits + 1), " ",
    ifelse(out$p.value < 0.05, "*", ""),
    ifelse(out$p.value < 0.01, "*", ""),
    ifelse(out$p.value < 0.001, "*", "")
  )
  
  out$ci <- paste0("(", out$conf.low.1, " - ", out$conf.high.1, ")")
  
  # Reference handling
  out$estimate.1[is.na(out$estimate)] <- refLabel
  out$stars[out$p.value < 0.001] <- "<0.001 ***"
  out$stars[is.na(out$estimate)] <- ""
  out$ci[is.na(out$estimate)] <- ""
  out$estimate[is.na(out$estimate)] <- 0
  
  # Clean display
  out$var <- as.character(out$var)
  out$var[duplicated(out$var)] <- ""
  out$N <- paste0("(N=", out$N, ")")
  
  # Reverse order (as in original)
  out <- out[nrow(out):1, ]
  
  return(out)
}
  


plot_multivariate_forest <- function(surv_dat, event_type = "PFS", max_x = 8) {
  
  # Select Event Type
  if(event_type == "PFS") {
    surv_dat <- surv_dat %>% 
      mutate(event = PFS_event, time = PFS_time)
  } else if(event_type == "OS") {
    surv_dat <- surv_dat %>% 
      mutate(event = OS_event, time = OS_time)
  } else {
    errorCondition("No event_type!")
  }
  
  # Select Features
  surv_dat <- surv_dat %>% 
    select(pat_id, score, event, time, 
           response_CR_d180, sex, age, admission_LDH, admission_ECOG, disease_cat, 
           antibiotics_pretreatment_cat, response_CRS_grade, sample_time,
           CART_costim, clinic_id)
  
  # Clean Features
  surv_dat_clean <- surv_dat %>%
    mutate(
      Sex = ifelse(sex == "w", 0,1),
      LDH = scale(admission_LDH),
      Age = scale(age),
      clinic_id = factor(clinic_id, labels = c("Heidelberg", "Regensburg", "München")),
      Center = fct_relevel(clinic_id, "Regensburg", "Heidelberg", "München"),
      ECOG = admission_ECOG,
      Antibiotics = antibiotics_pretreatment_cat,
      CRS_grade = response_CRS_grade,
      Disease = case_when(#disease_cat == "ALL" ~ "ALL",
        disease_cat %in% c("DLBCL", "FL", "MCL") ~ "NHL",
        disease_cat == "MM" ~ "MM",
        TRUE ~ "other/solid"),
      Disease = factor(Disease, levels = c("NHL", 
                                           "MM", 
                                           # "ALL",
                                           "other/solid")))
  
  # Fit Regression
  coxfit <- coxph(Surv(time, event) ~ Center + CART_costim + Disease + Antibiotics + CRS_grade + sample_time +
                    Sex + Age + LDH + ECOG + score, 
                  data = surv_dat_clean)
  
  # Clean Results
  clean_fit <- format_cox_table(coxfit, surv_dat_clean) %>% 
    mutate(var = ifelse(var == "score", "Risk Score\n(per pt.)", var),
           var = ifelse(var == "sample_time", "Sample Timing\n(Days to Therapy)", var),
           var = ifelse(var == "Sex", "Sex (M)", var),)
  
  clean_fit %>%
    rownames_to_column("id") %>% 
    mutate(label = sprintf("       %.2f\n        (%.2f–%.2f)", exp(estimate), exp(conf.low), exp(conf.high)),
           pval = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
           pval = signif(p.value, digits = 3),
           estimate = exp(estimate),
           conf.low = exp(conf.low),
           conf.high = exp(conf.high),
           id = factor(id, levels=rownames(clean_fit)),
           add_label = paste0(level, " ", N),
           label = ifelse(estimate.1 == "reference", "        reference", label)) %>% 
    ggplot(aes(y = id)) +
    geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                  width = 0.15, 
                  orientation = "y") +
    # Points (HRs)
    geom_point(aes(x = estimate), pch = 15, size = 2, color = "red") +
    # Vertical line at HR = 1
    geom_vline(xintercept = 1, linetype = "dashed") +
    # Add text labels on the right
    geom_text(aes(x = max_x * 4 , label = label), vjust = 0.5, size = 7/.pt) +
    geom_text(aes(x = max_x * 15 , label = ifelse(is.na(pval),"", paste("p =", pval))), hjust = 0, size = 7/.pt) +
    geom_text(aes(x = 0.02 , label = add_label), vjust = 0.5, hjust = 0, size = 7/.pt) +
    theme_prism(base_size = 7) +
    theme(axis.title.x = element_text(hjust = 0.4),
          axis.text.y = element_text(hjust=0, margin = margin(l = 20))) +
    # Axis & scales
    labs(
      y = NULL,
      x = "Hazard Ratio",
      title = NULL,
      caption = NULL
    ) +
    scale_x_log10(breaks = c(0.5,1,2,4,8)) +
    scale_y_discrete(labels = clean_fit$var) +
    coord_cartesian(clip = "off", xlim = c(0.02,max_x*50 )) +
    rremove("x.axis") +
    rremove("y.axis") +
    rremove("y.ticks")
}


```

### Risk Score
```{r}
metabolites <- c("Valeric acid",
                 "Isovaleric acid",
                 "Indole-3-acetic acid",
                 "Indole-3-carboxyaldehyd")

risk_score_df <- metabolites %>% 
  map_df(get_metabolite_group, data = metabo_data_stool) %>% 
  select(-cutpoint) %>% 
  mutate(value = as.character(value)) %>%
  mutate(value = case_when((name == "Valeric acid") & (value == "high") ~ 0,
                           (name == "Valeric acid") & (value == "low") ~ 1,
                           (name == "Isovaleric acid") & (value == "high") ~ 1,
                           (name == "Isovaleric acid") & (value == "low") ~ 0,
                           (name == "Indole-3-acetic acid") & (value == "high") ~ 1,
                           (name == "Indole-3-acetic acid") & (value == "low") ~ 0,
                           (name == "Indole-3-carboxyaldehyd") & (value == "high") ~ 1,
                           (name == "Indole-3-carboxyaldehyd") & (value == "low") ~ 0)) %>%
  pivot_wider() %>% 
  mutate(score = `Valeric acid` + `Isovaleric acid`  + `Indole-3-acetic acid` + `Indole-3-carboxyaldehyd`) %>% 
  select(pat_id, score)

p1 <- plt_risk_score(risk_score_df, clinic_data_stool, "PFS", binary_score = TRUE)

saveRDS(p1, file = paste0(fig6dir, "/01_RiskScoreBinaryFigure.rds"))

p2 <- plt_risk_score(risk_score_df, clinic_data_stool, "OS", binary_score = TRUE)

saveRDS(p2, file = paste0(fig6dir, "/02_RiskScoreBinaryFigure_OS.rds"))
```


### Multivariate Risk 
```{r}
# Forest Plots
p1 <- plot_multivariate_forest(risk_score_df %>% 
                                 left_join(clinic_data_stool),
                               event_type = "PFS")
p2 <- plot_multivariate_forest(risk_score_df %>% 
                                 left_join(clinic_data_stool),
                               event_type = "OS")

p1
saveRDS(p1, file = paste0(fig6dir, "/03_RiskScorePFSForest.rds"))
p2
saveRDS(p2, file = paste0(fig6dir, "/04_RiskScoreOSForest.rds"))
```

## Figure S13 - Stratified Risk Score
### Set Paths and Load Data

```{r}
figS13dir <- "../2_analysis/Figures/FigS13"
dir.create(figS13dir, recursive = T)
```

### Stratified for Centers
```{r}

clinic1_pat <- clinic_data_stool %>% 
  filter(clinic_id == 1) %>% 
  pull(pat_id)

risk_score_df_clinic1<- risk_score_df %>% filter(pat_id %in% clinic1_pat)

p1 <- plt_risk_score(risk_score_df_clinic1, clinic_data_stool, "PFS", binary_score = TRUE)
p2 <- plt_risk_score(risk_score_df_clinic1, clinic_data_stool, "OS", binary_score = TRUE)

clinic2_pat <- clinic_data_stool %>% 
  filter(clinic_id == 2) %>% 
  pull(pat_id)

risk_score_df_clinic2<- risk_score_df %>% filter(pat_id %in% clinic2_pat)

p3 <- plt_risk_score(risk_score_df_clinic2, clinic_data_stool, "PFS", binary_score = TRUE)
p4 <- plt_risk_score(risk_score_df_clinic2, clinic_data_stool, "OS", binary_score = TRUE)

clinic3_pat <- clinic_data_stool %>% 
  filter(clinic_id == 3) %>% 
  pull(pat_id)

risk_score_df_clinic3<- risk_score_df %>% filter(pat_id %in% clinic3_pat)

p5 <- plt_risk_score(risk_score_df_clinic3, clinic_data_stool, "PFS", binary_score = TRUE)
p6 <- plt_risk_score(risk_score_df_clinic3, clinic_data_stool, "OS", binary_score = TRUE)


saveRDS(list("HB_PFS" = p1,
             "HB_OS" = p2,
             "RB_PFS" = p3,
             "RB_OS" = p4,
             "MU_PFS" = p5,
             "MU_OS" = p6), file = paste0(figS13dir, "/01_RiskScoreBinaryCenter.rds"))

```


### Stratified for Disease Types

```{r}
# CD19 treated
CD19_pat <- clinic_data_stool %>% 
  filter(disease_cat %in% c("DLBCL", "FL", "MCL", "ALL")) %>% 
  pull(pat_id)

risk_score_df_cd19<- risk_score_df %>% filter(pat_id %in% CD19_pat)

p1 <- plt_risk_score(risk_score_df_cd19, clinic_data_stool, "PFS", binary_score = TRUE)
p2 <- plt_risk_score(risk_score_df_cd19, clinic_data_stool, "OS", binary_score = TRUE)

p3 <- ggarrange(p1, p2, ncol = 2)

# CD19+ NHL 
NHL_pat <- clinic_data_stool %>% 
  filter(disease_cat %in% c("DLBCL", "FL", "MCL")) %>% 
  pull(pat_id)

risk_score_df_nhl<- risk_score_df %>% filter(pat_id %in% NHL_pat)

p4 <- plt_risk_score(risk_score_df_nhl, clinic_data_stool, "PFS", binary_score = TRUE)
p5 <- plt_risk_score(risk_score_df_nhl, clinic_data_stool, "OS", binary_score = TRUE)

p6 <- ggarrange(p4, p5, ncol = 2)

# DLBCL
DLBCL_pat <- clinic_data_stool %>% 
  filter(disease_cat == "DLBCL") %>% 
  pull(pat_id)

risk_score_df_dlbcl <- risk_score_df %>% filter(pat_id %in% DLBCL_pat)

p7 <- plt_risk_score(risk_score_df_dlbcl, clinic_data_stool, "PFS", binary_score = TRUE)
p8 <- plt_risk_score(risk_score_df_dlbcl, clinic_data_stool, "OS", binary_score = TRUE)

p9 <- ggarrange(p7, p8, ncol = 2)

# MM
MM_pat <- clinic_data_stool %>% 
  filter(disease_cat == "MM") %>% 
  pull(pat_id)

risk_score_df_mm<- risk_score_df %>% filter(pat_id %in% MM_pat)

p10 <- plt_risk_score(risk_score_df_mm, clinic_data_stool, "PFS", binary_score = TRUE)
p11 <- plt_risk_score(risk_score_df_mm, clinic_data_stool, "OS", binary_score = TRUE)

p12 <- ggarrange(p10,p11, ncol = 2)

saveRDS(list("CD19" = p3,
             "CD19+ NHL" = p6,
             "DLBCL" = p9,
             "MM" = p12), file = paste0(figS13dir, "/02_RiskScoreBinaryDiseasetype.rds"))


```


## Figure S14 - Serum Risk Score
### Set Paths and Load Data 
```{r}
figS14dir <- "../2_analysis/Figures/FigS14"
dir.create(figS14dir, recursive = T)
```


### Risk Score Serum 

```{r}
metabolites <- c("Valeric acid",
                 "Isovaleric acid",
                 "Indole-3-acetic acid"
                 # "Indole-3-carboxyaldehyd",
                 )

risk_score_df_serum <- metabolites %>% 
  map_df(get_metabolite_group, data = metabo_data_serum) %>% 
  select(-cutpoint) %>% 
  mutate(value = as.character(value)) %>%
  mutate(value = case_when((name == "Valeric acid") & (value == "high") ~ 0,
                           (name == "Valeric acid") & (value == "low") ~ 1,
                           (name == "Isovaleric acid") & (value == "high") ~ 1,
                           (name == "Isovaleric acid") & (value == "low") ~ 0,
                           (name == "Indole-3-acetic acid") & (value == "high") ~ 1,
                           (name == "Indole-3-acetic acid") & (value == "low") ~ 0,
                           # (name == "Indole-3-carboxyaldehyd") & (value == "high") ~ 1,
                           # (name == "Indole-3-carboxyaldehyd") & (value == "low") ~ 0),
  )
  ) %>%
  pivot_wider() %>% 
  mutate(score = `Valeric acid` 
         + `Isovaleric acid`  
         + `Indole-3-acetic acid`
         # + `Indole-3-carboxyaldehyd`
         ) %>% 
  select(pat_id, score)

p1 <- plt_risk_score(risk_score_df_serum, clinic_data_serum, "PFS", binary_score = TRUE)

saveRDS(p1, file = paste0(figS14dir, "/01_RiskScoreBinaryFigure_Serum.rds"))

p2 <- plt_risk_score(risk_score_df_serum, clinic_data_serum, "OS", binary_score = TRUE)

saveRDS(p2, file = paste0(figS14dir, "/02_RiskScoreBinaryFigure_OS_Serum.rds"))


# Forste Plots of Multivariate Analysis
p1 <- plot_multivariate_forest(risk_score_df_serum %>% 
                                 left_join(clinic_data_serum),
                               event_type = "PFS")
p2 <- plot_multivariate_forest(risk_score_df_serum %>% 
                                 left_join(clinic_data_serum),
                               event_type = "OS")

p1
saveRDS(p1, file = paste0(figS14dir, "/03_RiskScorePFSForest_Serum.rds"))

p2
saveRDS(p2, file = paste0(figS14dir, "/04_RiskScoreOSForest_Serum.rds"))

```

### ROC and PR Curves
```{r}
# ROC
score_df <- risk_score_df %>% 
  left_join(clinic_data_stool %>% select(pat_id, PFS_event))

roc_res <- roc(score_df$PFS_event, score_df$score)
auc_res <- pROC::auc(score_df$PFS_event, score_df$score)

plt_roc <- pROC::ggroc(roc_res, size = 1.2, color = "black", legacy.axes = TRUE) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(subtitle = paste0("AUC = ", signif(auc_res, 3))) +
  theme_prism(base_size = 9, border = T) +
  coord_equal()

plt_roc
saveRDS(plt_roc, file = paste0(figS14dir, "/05_RiskScoreROC.rds"))

# PR Curve
df <- data.frame(
  actual_PFSevent = score_df$PFS_event,
  predicted_PFSevent = score_df$score
)

curves <- precrec::evalmod(scores = df$predicted_PFSevent, labels = df$actual_PFSevent)
prauc <- precrec::auc(curves) %>% pull(aucs) %>% .[2]
plt_pr <- autoplot(curves, "PRC") +
  theme_prism(base_size = 9) +
  labs(subtitle = paste0("AUC = ", signif(prauc, 3)),
       title = NULL) +
  theme(legend.position = "none")

plt_pr
saveRDS(plt_pr, file = paste0(figS14dir, "/06_RiskScorePR.rds"))


score_df_serum <- risk_score_df_serum %>% 
  left_join(clinic_data_serum %>% select(pat_id, PFS_event))

roc_res <- roc(score_df_serum$PFS_event, score_df_serum$score)
auc_res <- pROC::auc(score_df_serum$PFS_event, score_df_serum$score)

plt_roc <- pROC::ggroc(roc_res, size = 1.2, color = "black", legacy.axes = TRUE) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(subtitle = paste0("AUC = ", signif(auc_res, 3))) +
  theme_prism(base_size = 9, border = T) +
  coord_equal()

plt_roc
saveRDS(plt_roc, file = paste0(figS14dir, "/07_RiskScoreROC_Serum.rds"))

# PR Curve
df <- data.frame(
  actual_PFSevent = score_df_serum$PFS_event,
  predicted_PFSevent = score_df_serum$score
)

curves <- precrec::evalmod(scores = df$predicted_PFSevent, labels = df$actual_PFSevent)
prauc <- precrec::auc(curves) %>% pull(aucs) %>% .[2]
plt_pr <- autoplot(curves, "PRC") +
  theme_prism(base_size = 9) +
  labs(subtitle = paste0("AUC = ", signif(prauc, 3)),
       title = NULL) +
  theme(legend.position = "none")

plt_pr
saveRDS(plt_pr, file = paste0(figS14dir, "/08_RiskScorePR_Serum.rds"))


```

